---
format: pptx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(dplyr.summarise.inform = FALSE)
options(scipen = 999)

library(tidyverse)
library(broom)
library(readxl)
library(ggplot2)
library(plotly)
library(leaflet)
library(sf)
library(geosphere)
library(knitr)
library(scales)
library(kableExtra)
library(sp)
library(stats)
library(openxlsx)
library(gt)
library(ggrepel)
library(ggtext)
library(showtext)
showtext_auto()
```

```{r read_data, message=FALSE}
# -------------- 0.1.Import Files ---------------
# Read Data
HH <- read.csv("C:/Work/Travel Trends/HTS/2024 HTS/processed/hh_processed.csv", header = T, na.strings=c("NA","N/A", ""), colClasses=c("year"="character"))     %>% mutate(year = as.integer(year))
Person <- read.csv("C:/Work/Travel Trends/HTS/2024 HTS/processed/per_processed.csv", header = T, na.strings=c("NA","N/A", ""), colClasses=c("year"="character")) %>% mutate(year = as.integer(year))
Trips <- read.csv("C:/Work/Travel Trends/HTS/2024 HTS/processed/trips_processed.csv", header = T, na.strings=c("NA","N/A", ""), colClasses=c("year"="character")) %>% mutate(year = as.integer(year))
Day <- read.csv("C:/Work/Travel Trends/HTS/2024 HTS/processed/day_processed.csv", header = T, na.strings=c("NA","N/A", ""), colClasses=c("year"="character")) %>% mutate(year = as.integer(year))

# Wolfpack palette (ordered)
wolfpack_colors <- c(
  "#cc0000", # Wolfpack & Reynolds Red
  "#e16408", # Pyroman Flame
  "#fac800", # Hunt Yellow
  "#8d9e2b", # Genomic Green
  "#2db597", # Carmichael Aqua
  "#427e93", # Innovation Blue
  "#5b73bb"  # Bio-Indigo
)

# Discrete color/fill scales that recycle colors as needed
scale_color_wolfpack <- function(...) {
  ggplot2::discrete_scale("colour", "wolfpack",
    palette = scales::manual_pal(values = wolfpack_colors), ...
  )
}

scale_fill_wolfpack <- function(...) {
  ggplot2::discrete_scale("fill", "wolfpack",
    palette = scales::manual_pal(values = wolfpack_colors), ...
  )
}

# Helper: generate "nice" y-axis limits (and optional breaks) for any numeric vector.
nice_y_axis <- function(x,
                        n_breaks = 5,          # user desired number of ticks (labels)
                        max_breaks = 6,        # hard cap
                        allow_half = TRUE,     # allow half-step offset (e.g. 3.5, 4.5, ...)
                        nice_multipliers = c(1, 2, 2.5, 5, 10),
                        start_at_zero = FALSE  # if TRUE, force lower limit and first break at 0
                        ) {
  x <- x[is.finite(x)]
  if (!length(x)) {
    brks <- 0:1
    return(list(limits = range(brks), breaks = brks, step = 1, n_actual = length(brks)))
  }
  xmin <- min(x); xmax <- max(x)
  if (start_at_zero) xmin <- 0  # anchor lower bound request
  if (xmin == xmax) {
    mag  <- 10^floor(log10(abs(ifelse(xmin == 0, 1, xmin))))
    step <- mag
    lower <- xmin - step
    upper <- xmax + step
    brks  <- seq(lower, upper, by = step)
    return(list(limits = c(lower, upper), breaks = brks, step = step, n_actual = length(brks)))
  }

  # Cap requested breaks
  req <- max(2, min(n_breaks, max_breaks))
  rng <- xmax - xmin
  raw_step <- rng / max(1, (req - 1))
  mag <- 10^floor(log10(raw_step))

  # Candidate steps (include one order smaller / larger)
  candidates <- sort(unique(c(
    (mag/10) * nice_multipliers,
    mag      * nice_multipliers,
    (mag*10) * nice_multipliers
  )))
  candidates <- candidates[candidates > 0]

  best <- NULL
  score_best <- Inf

  for (st in candidates) {
    offsets <- c(0)
    if (allow_half) offsets <- unique(c(offsets, st/2))
    for (off in offsets) {
      # Shift logic
      # Compute an anchor-adjusted lower/upper so breaks align at lower = k*st + off
      lower <- floor((xmin - off) / st) * st + off
      # If user wants zero baseline, enforce lower <= 0 and adjust to exact 0 if within half step
      if (start_at_zero) {
        # Force non-negative lower bound
        if (lower < 0 && lower > -st*0.25) {
          # small negative due to rounding: snap to 0
          lower <- 0
        } else if (lower < 0) {
          # If substantially below 0, raise to 0 (do NOT extend below 0)
          lower <- 0
        } else if (lower > 0) {
          # If rounding placed us above 0, step back to 0
          lower <- 0
        }
      }
      upper <- ceiling((xmax - off) / st) * st + off
      if (lower >= xmin) lower <- lower - st
      if (upper <= xmax) upper <- upper + st

      # Re-assert zero if requested (the adjustment above can be offset by the generic expansion lines)
      if (start_at_zero) {
        if (lower < 0) lower <- 0
      }

      brks <- seq(lower, upper, by = st)

      # If too many breaks, skip (cannot exceed cap)
      if (length(brks) > max_breaks) next

      # If too few, extend upward (and downward if helpful) but do not exceed cap
      while (length(brks) < req && length(brks) < max_breaks) {
        upper <- upper + st
        brks  <- seq(lower, upper, by = st)
        if (length(brks) > max_breaks) {
          upper <- upper - st
          brks  <- seq(lower, upper, by = st)
          break
        }
      }

      # Ensure 0 is included if requested
      if (start_at_zero) {
        if (!any(abs(brks) < .Machine$double.eps^0.25)) {
          # Prepend 0 if within limits
          if (0 >= min(brks) && 0 <= max(brks)) {
            brks <- sort(unique(c(0, brks)))
          } else if (0 < min(brks)) {
            # Add 0 and possibly trim highest break if exceeding max_breaks
            brks <- sort(unique(c(0, brks)))
          }
        }
        # If adding 0 caused overflow, trim farthest break while retaining 0
        if (length(brks) > max_breaks) {
          # Remove either the highest or lowest (not zero) to get back to cap
            removable <- setdiff(brks, 0)
            # Prefer removing one farthest from data upper bound if both sides available
            if (length(removable)) {
              # Choose candidate whose removal keeps range covering xmax
              # simple heuristic: drop the largest if upper extension is large
              brks <- sort(head(brks, max_breaks))
            }
        }
        # Adjust limits to respect zero baseline
        if (min(brks) < 0) {
          # Remove negative breaks (should not exist if logic above worked) and re-add 0
          brks <- sort(unique(c(0, brks[brks >= 0])))
        }
        if (min(brks) > 0) brks <- sort(unique(c(0, brks)))
      }

      # Scoring:
      # 1) penalty for not meeting requested (prefer >= req)
      # 2) total expansion beyond data range
      n_actual <- length(brks)
      meets    <- as.integer(n_actual >= req)
      expansion <- (upper - lower) - rng
      # Lower score is better: prefer meeting req (meets==1), then fewer extra ticks above req, then minimal expansion
      score <- (meets == 0) * 1000 +                  # heavy penalty if not meeting
               abs(n_actual - req) * 50 +             # closeness to requested
               expansion / max(1, rng)                # relative expansion

      if (score < score_best) {
        score_best <- score
        best <- list(limits = c(lower, upper),
                     breaks = brks,
                     step = st,
                     n_actual = n_actual)
      }
    }
    # Early exit if we already hit exact request with minimal expansion
    if (!is.null(best) && best$n_actual == req) break
  }

  # Fallback (should not happen): simple pretty sequence
  if (is.null(best)) {
    st <- candidates[1]
    lower <- floor(xmin / st) * st
    if (start_at_zero) lower <- 0
    upper <- ceiling(xmax / st) * st
    brks  <- seq(lower, upper, by = st)
    if (length(brks) > max_breaks) {
      # Thin evenly
      idx <- unique(round(seq(1, length(brks), length.out = max_breaks)))
      brks <- brks[idx]
    }
    if (start_at_zero && !any(abs(brks) < .Machine$double.eps^0.25)) {
      brks[1] <- 0
    }
    if (start_at_zero) {
      brks <- brks[brks >= 0]
      if (length(brks) < 2) {
        # fabricate a second break using step heuristic
        st2 <- if (!is.null(st) && st > 0) st else max(1, xmax/5)
        brks <- c(0, st2)
      }
    }
    best <- list(limits = range(brks), breaks = brks, step = st, n_actual = length(brks))
  }

  best
}
```

# Triangle Travel Trend Analysis {background-image="C:/Git/TravelTrendAnalysis/script/2025/pictures/01.png"}

<span style="color: white;"><em>2016, 2018, 2021 and 2022 Household Travel Surveys</em></span>

::: {.pptx-footer style="text-align: center; font-size: 16pt; color: white;"}
Institute for Transportation Research and Education, <br>
North Carolina State University<br>  
September, 2025
:::

------------------------------------------------------------------------

# Table of Contents

-   01 Demographic
-   02 Household Travel
-   03 Person Travel
-   04 Other Travel Metrics

------------------------------------------------------------------------

# Triangle Recurring Household Travel Survey

Unweighted households, people and trips, after processing and cleaning.


```{r}
tab01 <- rbind(
  Households = sapply(sort(unique(HH$year)), function(y) length(unique(HH$hh_id[HH$year == y]))),
  Persons    = sapply(sort(unique(HH$year)), function(y) length(unique(Person$person_id[Person$year == y]))),
  Trips      = sapply(sort(unique(HH$year)), function(y) length(unique(Trips$trip_id[Trips$year == y])))
)
colnames(tab01) <- sort(unique(HH$year))

tab01 %>%
  as.data.frame() %>%
  tibble::rownames_to_column(" ") %>%
  gt()
```


------------------------------------------------------------------------

# Overall Trip Statistics

Unweighted households, people and trips, after processing and cleaning.

```{r}
tab02a <- Trips %>%
  mutate(
    per_trip = trip_weight,
    veh_trip = ifelse(grepl("auto", mode_simple), trip_weight, 0)
  ) %>%
  group_by(year) %>%
  summarise(
    "Total Person Trips" = sum(per_trip, na.rm = T),
    "Total Vehicle Trips" = sum(veh_trip, na.rm = T),
    'Avg. Person Trip Length' = weighted.mean(distance_miles, per_trip, na.rm = T),
    'Avg. Vehicle Trip Length' = weighted.mean(distance_miles, veh_trip, na.rm = T)
  )

tab02b <- tab02a %>%
  select(c("year", "Total Person Trips", "Total Vehicle Trips")) %>%
  pivot_longer(-year, names_to = "Trip Type", values_to = "Total Trips") %>%
  mutate(
    name_lab = if_else(year == 2024, `Trip Type`, NA_character_)
  )

tab02c <- tab02a %>%
  select(c("year", "Avg. Person Trip Length", "Avg. Vehicle Trip Length")) %>%
  pivot_longer(-year, names_to = "Trip Type", values_to = "Trip Length") %>%
  mutate(
    name_lab = if_else(year == 2024, `Trip Type`, NA_character_)
  )

tab02d <- HH %>%
  group_by(year) %>%
  summarise(
    hh = sum(hh_weight))

tab02e <- Person %>%
  group_by(year) %>%
  summarise(
    people = sum(person_weight, na.rm = T)) 

tab02f <- tab02a %>%
  left_join(tab02d, by = "year") %>%
  left_join(tab02e, by = "year") %>%
  mutate('Avg. Daily Trips per Household' = `Total Person Trips`/ hh,
         'Avg. Daily Trips per Person' = `Total Person Trips`/ people) %>%
  select(-c(people, hh, "Total Person Trips", "Total Vehicle Trips", "Avg. Person Trip Length", "Avg. Vehicle Trip Length")) %>%
  pivot_longer(-year, names_to = "Trip Type", values_to = "Trip Rates") %>%
  mutate(
    name_lab = if_else(year == 2024, `Trip Type`, NA_character_)
  )

sample_size_t <-  Trips %>%
  group_by(year) %>%
  summarise(
    samplesize = n_distinct(trip_id), .groups = "drop"
  )

```


```{r total trips, fig.width=9, fig.height=3, fig.align='center'}
yaxis <- nice_y_axis(tab02b$`Total Trips`, max_breaks = 5)

plt1 <- ggplot(tab02b, aes(x = year, y = `Total Trips`, color = `Trip Type`, group = `Trip Type`)) +
  # manually add grid lines
  geom_vline(
    xintercept = seq(min(tab02b$year), max(tab02b$year), by = 1),
    color = "grey91", 
    size = .6
  ) +
  geom_segment(
    data = tibble(y = yaxis$breaks, x1 = 2016, x2 = 2024),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey91",
    size = .6
  ) +
  # add data line and point
  geom_line(size = 1) +
  geom_point(size = 2) +
  # add line legend at line ends
  geom_text_repel(
    aes(color = `Trip Type`, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = 0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  )  +
  # add numeric data labels. kept INSIDE x-axis limits
  geom_text_repel(
    aes(color = `Trip Type`, label = scales::comma(round(`Total Trips`, 0))),  # format labels
    size = 3,
    direction = "both",
    xlim = c(2016, 2024),
    box.padding = 0.25,
    point.padding = 0.15,
    segment.size = 0.3,
    nudge_x = 0.1
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
    breaks = seq(min(tab02b$year), max(tab02b$year), 1),
    limits = c(min(tab02b$year), max(tab02b$year)+2),  
    expand = c(0.02, 0.02)
  ) +
  scale_y_continuous(
    labels = scales::comma,
    breaks = yaxis$breaks,
    limits = yaxis$limits,
    expand = c(0.05, 0.05)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Total Trips",
    color = NULL,
    title = "Overall Trip Statistics"
  ) +
  guides(color = "none") +
  # Sample size labels (one per year) using sample_size data frame
  annotate(
    "text",
    x = sample_size_t$year,
    y = yaxis$limits[2],
    label = paste0("n=", scales::comma(sample_size_t$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.35,        # < 0.5 = slightly left of centered
                              margin = margin(b = 20)),  # small space below title),
    plot.margin = margin(30, 40, 20, 10),
    # Remove the grid lines that come with ggplot2 plots by default
    panel.grid = element_blank())
    
plt1
```



```{r trip length, fig.width=9, fig.height=3, fig.align='center'}
yaxis <- nice_y_axis(tab02c$`Trip Length`, max_breaks = 5)

plt2 <- ggplot(tab02c, aes(x = year, y = `Trip Length`, color = `Trip Type`, group = `Trip Type`)) +
  # manually add grid lines
  geom_vline(
    xintercept = seq(min(tab02c$year), max(tab02c$year), by = 1),
    color = "grey91", 
    size = .6
  ) +
  geom_segment(
    data = tibble(y = yaxis$breaks, x1 = 2016, x2 = 2024),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey91",
    size = .6
  ) +
  # add data line and point
  geom_line(size = 1) +
  geom_point(size = 2) +
  # add line legend at line ends
  geom_text_repel(
    aes(color = `Trip Type`, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = 0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  )  +
  # add numeric data labels. kept INSIDE x-axis limits
  geom_text_repel(
    aes(color = `Trip Type`, label = round(`Trip Length`, 1)),  # format labels
    size = 3,
    direction = "both",
    xlim = c(2016, 2024),
    box.padding = 0.25,
    point.padding = 0.15,
    v_just = 1,
    nudge_y = 0.3,
    max.overlaps = Inf
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
    breaks = seq(min(tab02c$year), max(tab02c$year), 1),
    limits = c(min(tab02c$year), max(tab02c$year)+2),  
    expand = c(0.02, 0.02)
  ) +
  scale_y_continuous(
    labels = scales::comma,
    breaks = yaxis$breaks,
    limits = yaxis$limits,
    expand = c(0.05, 0.05)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Trip Length",
    color = NULL
  ) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(),
    plot.margin = margin(30, 40, 20, 10),
    # Remove the grid lines that come with ggplot2 plots by default
    panel.grid = element_blank())
    
plt2
```


```{r trip rate, fig.width=9, fig.height=3, fig.align='center'}
yaxis <- nice_y_axis(tab02f$`Trip Rates`, max_breaks = 5)

plt3 <- ggplot(tab02f, aes(x = year, y = `Trip Rates`, color = `Trip Type`, group = `Trip Type`)) +
  # manually add grid lines
  geom_vline(
    xintercept = seq(min(tab02f$year), max(tab02f$year), by = 1),
    color = "grey91", 
    size = .6
  ) +
  geom_segment(
    data = tibble(y = yaxis$breaks, x1 = 2016, x2 = 2024),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey91",
    size = .6
  ) +
  # add data line and point
  geom_line(size = 1) +
  geom_point(size = 2) +
  # add line legend at line ends
  geom_text_repel(
    aes(color = `Trip Type`, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2027, 2027),
    hjust = 0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  )  +
  # add numeric data labels. kept INSIDE x-axis limits
  geom_text_repel(
    aes(color = `Trip Type`, label = round(`Trip Rates`, 1)),  # format labels
    size = 3,
    direction = "both",
    xlim = c(2016, 2024),
    box.padding = 0.25,
    point.padding = 0.15,
    nudge_y = 0.3,
    v_just = 1,
    max.overlaps = Inf
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
    breaks = seq(min(tab02f$year), max(tab02f$year), 1),
    limits = c(min(tab02f$year), max(tab02f$year)+2),  
    expand = c(0.02, 0.02)
  ) +
  scale_y_continuous(
    labels = scales::comma,
    breaks = yaxis$breaks,
    limits = yaxis$limits,
    expand = c(0.05, 0.05)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Trip Rates",
    color = NULL
  ) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(),
    plot.margin = margin(30, 40, 20, 10),
    # Remove the grid lines that come with ggplot2 plots by default
    panel.grid = element_blank())
    
plt3
```

------------------------------------------------------------------------

# Triangle Demographics

Average household metrics have remained stable across the years.

```{r 1.5}
tab03a <- HH %>%
  group_by(year) %>%
  summarise(
    hh = sum(hh_weight),
    vehicle = sum(num_vehicles*hh_weight)
  ) %>%
  mutate(
    'Vehicles per Household' = vehicle/hh
  )

tab03b <- Person %>%
  mutate(
    driver = ifelse(can_drive== "Yes, drives", 1,0),
    worker = ifelse(num_jobs!="", 1,0)
  )%>%
  group_by(year) %>%
  summarise(
    people = sum(person_weight, na.rm = T),
    driver = sum(driver*person_weight, na.rm = T),
    worker = sum(worker*person_weight, na.rm = T)
  ) 

tab03 <- merge(tab03a, tab03b, by = "year")%>%
  mutate(
    'Person per Household' = people/hh,
    'Worker per Household' = worker/hh,
    'Driver per Household' = driver/hh,
    'Vehicles per Driver' = vehicle/driver,
    'Vehicles per Worker' = vehicle/worker
  ) %>%
  select(-hh, -vehicle, -people, -driver, -worker)%>%
  select(
    "year",
    'Person per Household',
    'Worker per Household',
    'Driver per Household',
    'Vehicles per Household',
    'Vehicles per Worker',
    'Vehicles per Driver'
  ) %>%
  mutate_if(is.numeric, format, digits=1, nsmall = 2) 

tab03_t <- tab03 %>%
  tibble::column_to_rownames("year") %>%
  t() 

tab03_t %>%
  as.data.frame() %>%
  tibble::rownames_to_column(" ") %>%
  gt()

```


------------------------------------------------------------------------

# Size & Workers

Household size is stable while the number of workers per household has shown an upward trend.

:::: {.columns}
::: {.column width="50%"}
```{r size}
sample_size_h <-  HH %>%
  group_by(year) %>%
  summarise(
    samplesize = n_distinct(hh_id), .groups = "drop"
  )

tab04a <- HH %>% 
  group_by(year, HHSize) %>%
  summarise(
    hhnum = sum(hh_weight)
  ) %>%
  mutate(
    Percentage = hhnum/sum(hhnum)
  ) %>%
  select(-hhnum) %>%
  mutate(
    HHSize = case_when(
        HHSize == "1" ~ "1 Person",
        HHSize == "2" ~ "2 People",
        HHSize == "3" ~ "3 People",
        HHSize == "4+" ~ "4+ People"
    ),
    name_lab = if_else(year == 2024, HHSize, NA_character_),
    # Ensure factor ordering low -> high so stacking (cumsum) adds 1,2,3,4+ in that order
    HHSize = factor(HHSize, levels = c("1 Person", "2 People", "3 People", "4+ People")))  %>%
  arrange(HHSize)
  


tab04a_labels <- tab04a %>%
  # HHSize already factored in the desired order; retain for clarity
  group_by(year) %>%
  arrange(HHSize, by_group = TRUE) %>%                        # stacking order (1 -> 4+)
  mutate(
    cum        = cumsum(Percentage),         # top boundary
    y_mid      = cum - Percentage / 2,       # middle of the band
    pct_label  = scales::percent(Percentage, accuracy = 1),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab04a, aes(x = year, y = Percentage, fill = HHSize)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  # middle labels
  geom_text(
    data = tab04a_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab04a_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab04a$year), max(tab04a$year), 1),
    limits = c(min(tab04a$year), max(tab04a$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE,
                    limits = c("1 Person", "2 People", "3 People", "4+ People")) +
  scale_color_wolfpack() +       
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Household Size"
  ) +
  # add line legend at line ends
  geom_text_repel(
    data = tab04a_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = HHSize, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = -0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  )  +
  # Sample size labels (one per year) using sample_size data frame
  annotate(
    "text",
    x = sample_size_h$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_h$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5,        # < 0.5 = slightly left of centered
                              margin = margin(b = 20)),  # small space below title),
    plot.margin = margin(30, 60, 20, 10)

  )


```
:::

::: {.column width="50%"}
```{r hhworkers}
tab04b <- HH %>%
  group_by(year, HHWorkers) %>%
  summarise(hhnum = sum(hh_weight), .groups = "drop_last") %>%
  mutate(Percentage = hhnum / sum(hhnum)) %>%
  select(-hhnum) %>%
  mutate(
    HHWorkers = case_when(
      HHWorkers == "0"  ~ "0 Workers",
      HHWorkers == "1"  ~ "1 Worker",
      HHWorkers == "2"  ~ "2 Workers",
      HHWorkers == "3+" ~ "3+ Workers"
    ),
    name_lab = if_else(year == 2024, HHWorkers, NA_character_),
    HHWorkers = factor(HHWorkers,
                            levels = c("0 Workers",  "1 Worker", "2 Workers","3+ Workers"))
  )

tab04b_labels <- tab04b %>%
  group_by(year) %>%
  arrange(HHWorkers, by_group = TRUE) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage / 2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab04b, aes(x = year, y = Percentage, fill = HHWorkers)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  # middle % labels inside bands
  geom_text(
    data = tab04b_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab04b_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab04b$year), max(tab04b$year), 1),
    limits = c(min(tab04b$year), max(tab04b$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Number of Workers"
  ) +
  # end labels (right side) for categories
  geom_text_repel(
    data = tab04b_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = HHWorkers, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(max(tab04b$year) + 0.5, NA),
    hjust = 0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  annotate(
    "text",
    x = sample_size_h$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_h$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 60, 20, 10)
  )
```
:::
::::

------------------------------------------------------------------------

# Auto Ownership & Income

Zero-vehicle households remain stable while households with more than 2 vehicles increased over time. Household income within the Triangle region is also increasing over the years.

:::: {.columns}
::: {.column width="50%"}
```{r auto ownership}
tab05a <- HH %>%
  group_by(year, nveh) %>%
  summarise(hhnum = sum(hh_weight), .groups = "drop_last") %>%
  mutate(Percentage = hhnum / sum(hhnum)) %>%
  select(-hhnum) %>%
  mutate(
    nveh = case_when(
      nveh == "0"  ~ "0 Vehicles",
      nveh == "1"  ~ "1 Vehicle",
      nveh == "2"  ~ "2 Vehicles",
      nveh == "3+" ~ "3+ Vehicles"
    ),
    name_lab = if_else(year == 2024, nveh, NA_character_)
  )

tab05a_labels <- tab05a %>%
  mutate(nveh = factor(nveh,
                       levels = c("0 Vehicles",  "1 Vehicle", "2 Vehicles", "3+ Vehicles"))) %>%
  group_by(year) %>%
  arrange(nveh) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage / 2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab05a, aes(x = year, y = Percentage, fill = nveh)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  geom_text(
    data = tab05a_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab05a_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab05a$year), max(tab05a$year), 1),
    limits = c(min(tab05a$year), max(tab05a$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Vehicle Ownership"
  ) +
  geom_text_repel(
    data = tab05a_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = nveh, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(max(tab05a$year) + 0.5, NA),
    hjust = 0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  annotate(
    "text",
    x = sample_size_h$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_h$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 60, 20, 10)
  )

```
:::
::: {.column width="50%"}
```{r hhincome}
tab05b <- HH %>%
  group_by(year, income_broad) %>%
  summarise(hhnum = sum(hh_weight), .groups = "drop_last") %>%
  mutate(Percentage = hhnum / sum(hhnum)) %>%
  select(-hhnum) %>%
  mutate(
    # Single factor definition: order from lowest income up
    income_broad = factor(
      income_broad,
      levels = c("Under $25,000", "$25,000-$49,999", "$50,000-$99,999", "$100,000 or more", "Prefer not to answer")
    ),
    name_lab = if_else(year == 2024, as.character(income_broad), NA_character_)
  )

tab05b_labels <- tab05b %>%
  group_by(year) %>%
  arrange(income_broad) %>%  # uses the single factor ordering defined above
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage / 2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab05b, aes(x = year, y = Percentage, fill = income_broad)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  geom_text(
    data = tab05b_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab05b_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(2016, 2024, 1),
    limits = c(2016, 2024),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Income Group"
  ) +
  geom_text_repel(
    data = tab05b_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = income_broad, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2024, 2026.5),
    hjust = 0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  annotate(
    "text",
    x = sample_size_h$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_h$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 110, 20, 10)
  )
```
:::

::::
------------------------------------------------------------------------

## Workers by Auto Ownership

The percentage of households with a given number of workers and vehicles has remained fairly stable in the past 5 years. 

```{r}
tab06a_tot <- HH %>%
  group_by(year) %>%
  summarise(hh_tot = sum(hh_weight))

tab06a <- HH %>%
  group_by(year, num_workersalt, nveh) %>%
  summarise(
    hh = sum(hh_weight)
  ) %>%
  left_join(tab06a_tot, by = "year") %>%
  mutate(
    Percentage = hh/hh_tot
  ) %>%
  select(-hh, -hh_tot) %>%
  pivot_wider(names_from = "num_workersalt", values_from = Percentage)

tab06a_t <- tab06a %>%
  t()  %>%
  as.data.frame() 

rows_to_format <- c("0", "1", "2", "3+")

tab06a_t[rownames(tab06a_t) %in% rows_to_format, ] <- 
  lapply(tab06a_t[rownames(tab06a_t) %in% rows_to_format, ], function(x) {
    # Convert to numeric, handle NAs, and format as percent with 1 decimal
    scales::percent(as.numeric(x), accuracy = 0.1)
  })

tab06a_t %>%
  tibble::rownames_to_column(" ") %>%
  gt()


```

------------------------------------------------------------------------

## Age & Gender

The distribution of survey data by age shows some lumpiness in the various age categories, likely attributed to small sample sizes and the fact that age is not a control factor in the data weighting and expansion. Even so, seniors are a growing segment of the Triangle’s population. Also, in 2018, survey data became more inclusive with new non-binary gender categories.

:::: {.columns}
::: {.column width="50%"}
```{r age}
sample_size_p <-  Person %>%
  group_by(year) %>%
  summarise(
    samplesize = n_distinct(person_id), .groups = "drop"
  )

# Age distribution stacked area (pattern mirrors HHSize chunk tab04a)
tab07a <- Person %>%
  group_by(year, agecat) %>%
  summarise(pop = sum(person_weight), .groups = "drop_last") %>%
  mutate(Percentage = pop / sum(pop)) %>%
  select(-pop) %>%
  mutate(
    # Factor order youngest -> oldest so cumulative stacking builds in that order
    agecat = factor(agecat, levels = c("Under 18", "18-34", "35-64", "65+")),
    name_lab = if_else(year == 2024, as.character(agecat), NA_character_)
  ) %>%
  arrange(agecat)

tab07a_labels <- tab07a %>%
  group_by(year) %>%
  arrange(agecat, by_group = TRUE) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage/2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab07a, aes(x = year, y = Percentage, fill = agecat)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  # internal percentage labels
  geom_text(
    data = tab07a_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab07a_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab07a$year), max(tab07a$year), 1),
    limits = c(min(tab07a$year), max(tab07a$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE,
                      limits = c("Under 18", "18-34", "35-64", "65+")) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Age"
  ) +
  # end labels only for final year (2024)
  geom_text_repel(
    data = tab07a_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = agecat, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = -0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  #sample size annotation: using person sample_size 
  annotate(
    "text",
    x = sample_size_p$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_p$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 60, 20, 10)
  )

```
:::
::: {.column width="50%"}
```{r gender}
# Gender distribution stacked area (pattern mirrors age chunk)

tab07b <- Person %>%
  group_by(year, gender) %>%
  summarise(pop = sum(person_weight), .groups = "drop_last") %>%
  mutate(Percentage = pop / sum(pop)) %>%
  select(-pop) %>%
  mutate(
    # Ensure consistent factor order (adjust if you prefer a different vertical ordering)
    gender = factor(gender),
    name_lab = if_else(year == 2024, as.character(gender), NA_character_)
  ) %>%
  arrange(gender)

tab07b_labels <- tab07b %>%
  group_by(year) %>%
  arrange(gender, by_group = TRUE) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage/2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab07b, aes(x = year, y = Percentage, fill = gender)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  geom_text(
    data = tab07b_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab07b_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab07b$year), max(tab07b$year), 1),
    limits = c(min(tab07b$year), max(tab07b$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Gender"
  ) +
  geom_text_repel(
    data = tab07b_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = gender, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = -0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  # sample size annotation (person-level, reuse sample_size_p if already defined like age chunk)
  annotate(
    "text",
    x = sample_size_p$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_p$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 60, 20, 10)
  )

```
:::
::::

------------------------------------------------------------------------

## Drivers

blank

:::: {.columns}
::: {.column width="50%"}
```{r drivers_age}
# Licensed drivers by age - stacked area (mirrors Age stacked pattern)

tab08a <- Person %>%
  filter(can_drive == "Yes, drives") %>%
  group_by(year, agecat) %>%
  summarise(pop = sum(person_weight), .groups = "drop_last") %>%
  mutate(Percentage = pop / sum(pop)) %>%
  select(-pop) %>%
  mutate(
    agecat  = factor(agecat, levels = c("Under 18", "18-34", "35-64", "65+")),
    name_lab = if_else(year == 2024, as.character(agecat), NA_character_)
  ) %>%
  arrange(agecat)

tab08a_labels <- tab08a %>%
  group_by(year) %>%
  arrange(agecat, by_group = TRUE) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage/2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab08a, aes(x = year, y = Percentage, fill = agecat)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  geom_text(
    data = tab08a_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab08a_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab08a$year), max(tab08a$year), 1),
    limits = c(min(tab08a$year), max(tab08a$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE,
                      limits = c("Under 18", "18-34", "35-64", "65+")) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Licensed Drivers by Age"
  ) +
  geom_text_repel(
    data = tab08a_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = agecat, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = -0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  # person sample sizes (re-use sample_size_p if defined earlier; else remove or define)
  annotate(
    "text",
    x = sample_size_p$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_p$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 60, 20, 10)
  )

```
:::
::: {.column width="50%"}
```{r drivers_gender}
# Licensed drivers by gender - stacked area (mirrors drivers_age pattern)

tab08b <- Person %>%
  filter(can_drive == "Yes, drives") %>%
  group_by(year, gender) %>%
  summarise(pop = sum(person_weight), .groups = "drop_last") %>%
  mutate(Percentage = pop / sum(pop)) %>%
  select(-pop) %>%
  mutate(
    gender  = factor(gender),  # set explicit levels if preferred
    name_lab = if_else(year == 2024, as.character(gender), NA_character_)
  ) %>%
  arrange(gender)

tab08b_labels <- tab08b %>%
  group_by(year) %>%
  arrange(gender, by_group = TRUE) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage/2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab08b, aes(x = year, y = Percentage, fill = gender)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  geom_text(
    data = tab08b_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab08b_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab08b$year), max(tab08b$year), 1),
    limits = c(min(tab08b$year), max(tab08b$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Licensed Drivers by Gender"
  ) +
  geom_text_repel(
    data = tab08b_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = gender, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = -0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  annotate(
    "text",
    x = sample_size_p$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_p$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 60, 20, 10)
  )

```
:::
::::

------------------------------------------------------------------------

## Workers

blank

:::: {.columns}
::: {.column width="50%"}
```{r workers_age}
# Workers by age - stacked area (mirrors Age chart pattern)

tab09a <- Person %>%
  filter(num_jobs != "") %>%
  group_by(year, agecat) %>%
  summarise(pop = sum(person_weight), .groups = "drop_last") %>%
  mutate(Percentage = pop / sum(pop)) %>%
  select(-pop) %>%
  mutate(
    agecat  = factor(agecat, levels = c("Under 18", "18-34", "35-64", "65+")),
    name_lab = if_else(year == 2024, as.character(agecat), NA_character_)
  ) %>%
  arrange(agecat)

tab09a_labels <- tab09a %>%
  group_by(year) %>%
  arrange(agecat, by_group = TRUE) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage/2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab09a, aes(x = year, y = Percentage, fill = agecat)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  geom_text(
    data = tab09a_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab09a_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab09a$year), max(tab09a$year), 1),
    limits = c(min(tab09a$year), max(tab09a$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE,
                      limits = c("Under 18", "18-34", "35-64", "65+")) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Workers by Age"
  ) +
  geom_text_repel(
    data = tab09a_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = agecat, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = -0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  annotate(
    "text",
    x = sample_size_p$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_p$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 60, 20, 10)
  )

```
:::
::: {.column width="50%"}
```{r workers_gender}
# Workers by gender - stacked area (mirrors workers_age pattern)

tab09b <- Person %>%
  filter(num_jobs != "") %>%
  group_by(year, gender) %>%
  summarise(pop = sum(person_weight), .groups = "drop_last") %>%
  mutate(Percentage = pop / sum(pop)) %>%
  select(-pop) %>%
  mutate(
    gender  = factor(gender),
    name_lab = if_else(year == 2024, as.character(gender), NA_character_)
  ) %>%
  arrange(gender)

tab09b_labels <- tab09b %>%
  group_by(year) %>%
  arrange(gender, by_group = TRUE) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage/2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    pct_label = ifelse(Percentage < 0.01, "", pct_label),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab09b, aes(x = year, y = Percentage, fill = gender)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  geom_text_repel(
    data = tab09b_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    box.padding = 0.4,
    min.segment.length = 0,
    nudge_x = tab09b_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab09b$year), max(tab09b$year), 1),
    limits = c(min(tab09b$year), max(tab09b$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Workers by Gender"
  ) +
  geom_text_repel(
    data = tab09b_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = gender, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2027, NA),
    hjust = -0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20,
    max.overlaps = 30
  ) +
  annotate(
    "text",
    x = sample_size_p$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_p$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 100, 20, 10)
  )

```
:::
::::

------------------------------------------------------------------------

## Average Daily Trips by Household

use pictogram here
The number of average daily trips made by households in the Triangle region is closer to rates observed pre-pandemic. 

```{r v trip rate}
# Average Daily Vehicle Trips per Household - Bar Chart

tab10a <- Trips %>%
  mutate(veh_trip = ifelse(grepl("auto", mode_type), trip_weight, 0)) %>%
  group_by(year) %>%
  summarise(tot_veh_trip = sum(veh_trip), .groups = "drop")

tab10b <- HH %>%
  group_by(year) %>%
  summarise(hh = sum(hh_weight), .groups = "drop")

tab10 <- tab10a %>%
  left_join(tab10b, by = "year") %>%
  mutate(`Daily vehicle trips/HH` = tot_veh_trip / hh) %>%
  select(year, `Daily vehicle trips/HH`)

yaxis <- nice_y_axis(tab10$`Daily vehicle trips/HH`, max_breaks = 6, start_at_zero = TRUE)
y_breaks <- yaxis$breaks
y_limits <- c(0, max(yaxis$breaks))


ggplot(tab10, aes(x = factor(year), y = `Daily vehicle trips/HH`, fill = factor(year))) +
  geom_col(width = 0.55) +
  geom_text(
    aes(label = round(`Daily vehicle trips/HH`, 1)),
    vjust = -0.6,
    size = 3.5,
    color = "grey20"
  ) +
  scale_y_continuous(breaks = y_breaks, limits = y_limits, expand = expansion(mult = c(0, 0.05))) +
  scale_fill_wolfpack() +
  labs(
    x = NULL,
    y = "Daily vehicle trips per household",
    title = "Average Daily Vehicle Trips per Household"
  ) +
  guides(fill = "none") +
  # Sample size labels below x-axis: place just below 0 using negative y offset
  annotate(
    "text",
    x = factor(sample_size_t$year),
    y = 0,
    label = paste0("n=", scales::comma(sample_size_t$samplesize)),
    color = "grey40",
    size = 3.2,
    vjust = 4
  ) +
  coord_cartesian(ylim=c(0,8), clip = "off") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 20, 40, 5)  # extra bottom room for annotation
  )

```



------------------------------------------------------------------------

## Trips by Mode of Travel

Post pandemic, the share of trips made by people traveling in autos has dropped as people are using alternative modes more than they did during the pandemic.

:::: {.columns}
::: {.column width="70%"}
```{r non auto ms}
# Non-Auto Mode Share - Stacked Area

tab11a <- Trips %>%
  filter(mode_simple != "auto") %>%
  group_by(year, mode_simple) %>%
  summarise(tot_per_trip = sum(trip_weight), .groups = "drop_last") %>%
  mutate(Percentage = tot_per_trip / sum(tot_per_trip)) %>%
  select(-tot_per_trip) %>%
  mutate(
    mode_simple = factor(mode_simple, levels = c("bus", "bike", "walk", "school bus", "other")),
    name_lab = if_else(year == 2024, as.character(mode_simple), NA_character_)
  ) %>%
  arrange(mode_simple)

tab11a_labels <- tab11a %>%
  group_by(year) %>%
  arrange(mode_simple, by_group = TRUE) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage/2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    pct_label = ifelse(Percentage < 0.01, "", pct_label),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab11a, aes(x = year, y = Percentage, fill = mode_simple)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  geom_text(
    data = tab11a_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    lineheight = 0.9,
    nudge_x = tab11a_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab11a$year), max(tab11a$year), 1),
    limits = c(min(tab11a$year), max(tab11a$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE,
                      limits = c("bus", "bike", "walk", "school bus", "other")) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Mode Share (Non-auto Modes Only)"
  ) +
  geom_text_repel(
    data = tab11a_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = mode_simple, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = -0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  annotate(
    "text",
    x = sample_size_t$year,
    y = 1,
    label = paste0("n=", scales::comma(sample_size_t$samplesize)),
    color = "grey40",
    size = 3.5,
    vjust = -1
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 60, 20, 10)
  )

```
:::
::: {.column width="30%"}
```{r ms}
# Mode Share (All Modes) - Stacked Area (patterned after non auto ms)

tab11b <- Trips %>%
  group_by(year, mode_simple) %>%
  summarise(tot_per_trip = sum(trip_weight), .groups = "drop_last") %>%
  mutate(Percentage = tot_per_trip / sum(tot_per_trip)) %>%
  select(-tot_per_trip) %>%
  mutate(
    # Include auto if present; reorder as desired
    mode_simple = factor(mode_simple, levels = c("auto","bus", "bike", "walk", "school bus", "other")),
    name_lab = if_else(year == 2024, as.character(mode_simple), NA_character_)
  ) %>%
  arrange(mode_simple)

tab11b_labels <- tab11b %>%
  group_by(year) %>%
  arrange(mode_simple, by_group = TRUE) %>%
  mutate(
    cum       = cumsum(Percentage),
    y_mid     = cum - Percentage/2,
    pct_label = scales::percent(Percentage, accuracy = 1),
    pct_label = ifelse(Percentage < 0.01, "", pct_label),
    nudgex = case_when(
      year == 2016 ~ 0.22,
      year == 2024 ~ -0.22,
      TRUE ~ 0
    )
  ) %>%
  ungroup()

ggplot(tab11b, aes(x = year, y = Percentage, fill = mode_simple)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9, position = position_fill(reverse = TRUE)) +
  geom_text_repel(
    data = tab11b_labels,
    aes(y = y_mid, label = pct_label),
    size = 3,
    show.legend = FALSE,
    vjust = 0.5,
    nudge_x = tab11b_labels$nudgex,
    color = "grey30"
  ) +
  scale_x_continuous(
    breaks = seq(min(tab11b$year), max(tab11b$year), 1),
    limits = c(min(tab11b$year), max(tab11b$year)),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack(drop = FALSE,
                      limits = c("auto","bus", "bike", "walk", "school bus", "other")) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Mode Share (All Modes)"
  ) +
  geom_text_repel(
    data = tab11b_labels %>% filter(!is.na(name_lab)),
    aes(y = y_mid, color = mode_simple, label = name_lab),
    fontface = "bold",
    size = 3.5,
    direction = "y",
    xlim = c(2025.5, NA),
    hjust = -0,
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted",
    box.padding = .4,
    segment.curvature = -0.1,
    segment.ncp = 3,
    segment.angle = 20
  ) +
  guides(fill = "none", color = "none") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5, margin = margin(b = 20)),
    plot.margin = margin(30, 60, 20, 10)
  )

```
:::
::::

------------------------------------------------------------------------

## Average Trip Lengths by Trip Purpose

Trip lengths post pandemic are returning to pre-pandemic levels and sometimes increasing. The most dramatic increase is for other to other trips.


```{r avg length by purpose}
tab12 <- Trips %>%
  group_by(year, trip_purpose) %>%
  summarise(
    distance_miles = weighted.mean(distance_miles,trip_weight,na.rm = T)) %>%
  filter(!is.na(trip_purpose))

tab12$trip_purpose <- factor(
  tab12$trip_purpose,
  levels = c("Home to Work","Home to Shop","Home to Other","Home to University","Home to School","Other to other")
)

# Create the line graph
ggplot(tab12, aes(x = year, y = distance_miles, color = trip_purpose, group = trip_purpose)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab12$year), max(tab12$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "distance_miles",
    color = NULL,
    title = "Average Trip Lengths by Trip Purpose"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Average Trip Duration by Trip Purpose

Trip lengths post pandemic are returning to pre-pandemic levels and sometimes increasing. The most dramatic increase is for other to other trips.


```{r}
tab31 <- Trips %>%
  group_by(year, trip_purpose) %>%
  summarise(
    duration_minutes = weighted.mean(duration_minutes,trip_weight,na.rm = T)) %>%
  filter(!is.na(trip_purpose))

tab31$trip_purpose <- factor(
  tab31$trip_purpose,
  levels = c("Home to Work","Home to Shop","Home to Other","Home to University","Home to School","Other to other")
)

# Create the line graph
ggplot(tab31, aes(x = year, y = duration_minutes, color = trip_purpose, group = trip_purpose)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab31$year), max(tab31$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "duration_minutes",
    color = NULL,
    title = "Average Trip Duration by Trip Purpose"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```


------------------------------------------------------------------------

## Average Trip Lengths by Mode

Trip lengths post pandemic are returning to pre-pandemic levels and sometimes increasing. The most dramatic increase is for other to other trips.


```{r}
tab32 <- Trips %>%
  group_by(year, mode_simple) %>%
  summarise(
    distance_miles = weighted.mean(distance_miles,trip_weight,na.rm = T)) %>%
  filter(!is.na(mode_simple) & mode_simple != "other")

tab32$mode_simple <- factor(
  tab32$mode_simple,
  levels = c("auto", "bus", "school bus", "bike", "walk")
)

# Create the line graph
ggplot(tab32, aes(x = year, y = distance_miles, color = mode_simple, group = mode_simple)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab32$year), max(tab32$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "distance_miles",
    color = NULL,
    title = "Average Trip Lengths by Mode"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Average Trip Duration by Mode

Trip lengths post pandemic are returning to pre-pandemic levels and sometimes increasing. The most dramatic increase is for other to other trips.


```{r}
tab33 <- Trips %>%
  group_by(year, mode_simple) %>%
  summarise(
    duration_minutes = weighted.mean(duration_minutes,trip_weight,na.rm = T)) %>%
  filter(!is.na(mode_simple) & mode_simple != "other")

tab33$mode_simple <- factor(
  tab33$mode_simple,
  levels = c("auto", "bus", "school bus", "bike", "walk")
)

# Create the line graph
ggplot(tab33, aes(x = year, y = duration_minutes, color = mode_simple, group = mode_simple)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab31$year), max(tab31$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "duration_minutes",
    color = NULL,
    title = "Average Trip Duration by Trip Purpose"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Reason for Travel

The reasons people travel remain fairly stable over time. Post pandemic, the average number of trips per day has increased almost back to original levels. The largest increase was for trips that neither began nor ended at home.

:::: {.columns}
::: {.column width="50%"}
```{r 13 trip_rate}
tab13a <- Person %>%
  group_by(year) %>%
  summarise(
    people = sum(person_weight),
  ) 

tab13b <- Trips %>%
  group_by(year, trip_purpose) %>%
  summarise(
    tot_per_trip = sum(trip_weight))

tab13 <- tab13b %>%
  left_join(tab13a, by = "year") %>%
  mutate(
    trip_rate = tot_per_trip/people
  )%>%
  select(-tot_per_trip, -people) %>%
  filter(!is.na(trip_purpose))

tab13$trip_purpose <- factor(
  tab13$trip_purpose,
  levels = c("Home to Work","Home to Shop","Home to Other","Home to University","Home to School","Other to other")
)

# Create the line graph
ggplot(tab13, aes(x = year, y = trip_rate, color = trip_purpose, group = trip_purpose)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab13$year), max(tab13$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "trip_rate",
    color = NULL,
    title = "Average Trip Rates by Trip Purpose"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```
:::
::: {.column width="50%"}
```{r 13 percentage}
# Percentage of Trips by Trip Purpose (line graph)
tab13c <- Trips %>%
  filter(!is.na(trip_purpose)) %>%
  group_by(year, trip_purpose) %>%
  summarise(tot_per_trip = sum(trip_weight), .groups = "drop_last") %>%
  mutate(Percentage = tot_per_trip / sum(tot_per_trip)) %>%
  ungroup() 

tab13c$trip_purpose <- factor(
  tab13c$trip_purpose,
  levels = c("Home to Work","Home to Shop","Home to Other","Home to University","Home to School","Other to other")
)

ggplot(tab13c, aes(x = year, y = Percentage, color = trip_purpose, group = trip_purpose)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab13c$year), max(tab13c$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage of Trips",
    color = NULL,
    title = "Percent of Trips by Trip Purpose"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

:::
::::

------------------------------------------------------------------------

## Travel by Income Groups

While the percentage of trips made by households with income greater than $100K has increased, the average number of trips per day is almost what it was pre-pandemic.

:::: {.columns}
::: {.column width="50%"}
```{r 14 trip_rate by income}
tab14a <- HH %>%
  group_by(year, income_broad) %>%
  summarise(
    household = sum(hh_weight)
  ) 

tab14b <- Trips %>%
  group_by(year, income_broad) %>%
  summarise(
    tot_per_trip = sum(trip_weight))

tab14 <- tab14b %>%
  left_join(tab14a, by = c("year", "income_broad")) %>%
  mutate(
    trip_rate = tot_per_trip/household
  )%>%
  select(-tot_per_trip, -household)

tab14$income_broad <- factor(
  tab14$income_broad,
  levels = c("Under $25,000", "$25,000-$49,999", "$50,000-$99,999", "$100,000 or more", "Prefer not to answer")
)

# Create the line graph
ggplot(tab14, aes(x = year, y = trip_rate, color = income_broad, group = income_broad)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab14$year), max(tab14$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "trip_rate",
    color = NULL,
    title = "Average Trip Rates by Household Income"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```
:::
::: {.column width="50%"}
```{r 14 percentage }
# Percentage of Trips by Income Group (line graph)
tab14c <- Trips %>%
  group_by(year, income_broad) %>%
  summarise(tot_per_trip = sum(trip_weight), .groups = "drop_last") %>%
  mutate(Percentage = tot_per_trip / sum(tot_per_trip)) %>%
  ungroup() 

tab14c$income_broad <- factor(
  tab14c$income_broad,
  levels = c("Under $25,000", "$25,000-$49,999", "$50,000-$99,999", "$100,000 or more", "Prefer not to answer")
)

ggplot(tab14c, aes(x = year, y = Percentage, color = income_broad, group = income_broad)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab14c$year), max(tab14c$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage of Trips",
    color = NULL,
    title = "Percent of Trips by Household Income"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

:::
::::


------------------------------------------------------------------------

## Travel by Household Size

Across all years, one-person households average slightly more trips on a per person basis. However, in terms of total trips, larger households make more trips due to the greater number of people in the household.


:::: {.columns}
::: {.column width="50%"}
```{r 15 trip_rate}
tab15a <- HH %>%
  group_by(year, HHSize) %>%
  summarise(
    household = sum(hh_weight)
  ) 

tab15b <- Trips %>%
  left_join(HH %>% select(c("hh_id", "HHSize")), by = "hh_id") %>%
  group_by(year, HHSize) %>%
  summarise(
    tot_per_trip = sum(trip_weight))

tab15 <- tab15b %>%
  left_join(tab15a, by = c("year", "HHSize")) %>%
  mutate(
    trip_rate = tot_per_trip/household
  )%>%
  select(-tot_per_trip, -household)

tab15$HHSize <- factor(tab15$HHSize, levels = c("1", "2", "3", "4+"), labels = c("1 Person", "2 People", "3 People", "4+ People"))

# Create the line graph
ggplot(tab15, aes(x = year, y = trip_rate, color = HHSize, group = HHSize)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab15$year), max(tab15$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "trip_rate",
    color = NULL,
    title = "Average Trip Rates by Household Size"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```
:::
::: {.column width="50%"}
```{r 15 percentage }
# Percentage of Trips by Income Group (line graph)
tab15c <- Trips %>%
  left_join(HH %>% select(c("hh_id", "HHSize")), by = "hh_id") %>%
  group_by(year, HHSize) %>%
  summarise(tot_per_trip = sum(trip_weight), .groups = "drop_last") %>%
  mutate(Percentage = tot_per_trip / sum(tot_per_trip)) %>%
  ungroup() 

tab15c$HHSize <- factor(tab15c$HHSize, levels = c("1", "2", "3", "4+"), labels = c("1 Person", "2 People", "3 People", "4+ People"))

ggplot(tab15c, aes(x = year, y = Percentage, color = HHSize, group = HHSize)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab15c$year), max(tab15c$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage of Trips",
    color = NULL,
    title = "Percent of Trips by Household Income"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

:::
::::

------------------------------------------------------------------------

## Travel by Household Workers

```{r trip rate by hh worker}
tab26a <- HH %>%
  group_by(year, num_workersalt) %>%
  summarise(
    hh = sum(hh_weight),
  ) 

tab26b <- Trips %>%
  left_join(HH %>% select(c("hh_id", "num_workersalt")), by = "hh_id") %>%
  group_by(year, num_workersalt) %>%
  summarise(
    tot_per_trip = sum(trip_weight))

tab26 <- tab26b %>%
  left_join(tab26a, by = c("year", "num_workersalt")) %>%
  mutate(
    trip_rate = tot_per_trip/hh
  )%>%
  select(-tot_per_trip, -hh) 


# Create the line graph
ggplot(tab26, aes(x = year, y = trip_rate, color = num_workersalt, group = num_workersalt)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab26b$year), max(tab26b$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "trip_rate",
    color = NULL,
    title = "Average Trip Rates by Number of Household Workers"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Travel by Auto Ownership

In the auto dominant Triangle region, households that do not own a vehicle have limited mobility in comparison to households with vehicles. Prior to the pandemic, auto ownership had the greatest impact on average trips for 1-auto households. As the number of autos in a household increases, total trips increase; however, the impact of the additional autos on trip making declines.


:::: {.columns}
::: {.column width="50%"}
```{r 16 trip_rate by nveh}
tab16a <- HH %>%
  group_by(year, nveh) %>%
  summarise(
    household = sum(hh_weight)
  ) 

tab16b <- Trips %>%
  left_join(HH %>% select(c("hh_id", "nveh")), by = "hh_id") %>%
  group_by(year, nveh) %>%
  summarise(
    tot_per_trip = sum(trip_weight))

tab16 <- tab16b %>%
  left_join(tab16a, by = c("year", "nveh")) %>%
  mutate(
    trip_rate = tot_per_trip/household
  )%>%
  select(-tot_per_trip, -household)

tab16$nveh <- factor(tab16$nveh, levels = c("0", "1", "2", "3+"), labels = c("0 Vehicle", "1 Vehicles",
  "2 Vehicles", "3+ Vehicles"))

# Create the line graph
ggplot(tab16, aes(x = year, y = trip_rate, color = nveh, group = nveh)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab16$year), max(tab16$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "trip_rate",
    color = NULL,
    title = "Average Trip Rates by Auto Ownership"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```
:::
::: {.column width="50%"}
```{r 16 percentage }
# Percentage of Trips by Income Group (line graph)
tab16c <- Trips %>%
  left_join(HH %>% select(c("hh_id", "nveh")), by = "hh_id") %>%
  group_by(year, nveh) %>%
  summarise(tot_per_trip = sum(trip_weight), .groups = "drop_last") %>%
  mutate(Percentage = tot_per_trip / sum(tot_per_trip)) %>%
  ungroup() 

tab16c$nveh <- factor(tab16c$nveh, levels = c("0", "1", "2", "3+"), labels = c("0 Vehicle", "1 Vehicles",
  "2 Vehicles", "3+ Vehicles"))

ggplot(tab16c, aes(x = year, y = Percentage, color = nveh, group = nveh)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab16c$year), max(tab16c$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage of Trips",
    color = NULL,
    title = "Percent of Trips by Auto Ownership"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

:::
::::

------------------------------------------------------------------------

## Person Travel

left blank


------------------------------------------------------------------------

## Average Daily Trips by Age Group

```{r trip rate by ager}
tab27a <- Person %>%
  group_by(year, agecat) %>%
  summarise(
    people = sum(person_weight),
  ) 

tab27b <- Trips %>%
  left_join(Person %>% select(c("person_id", "agecat")), by = "person_id") %>%
  group_by(year, agecat) %>%
  summarise(
    tot_per_trip = sum(trip_weight))

tab27 <- tab27b %>%
  left_join(tab27a, by = c("year", "agecat")) %>%
  mutate(
    trip_rate = tot_per_trip/people
  )%>%
  select(-tot_per_trip, -people) 

tab27$agecat <- factor(
  tab27$agecat,
  levels = c("Under 18", "18-34", "35-64", "65+")
)


# Create the line graph
ggplot(tab27, aes(x = year, y = trip_rate, color = agecat, group = agecat)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab26b$year), max(tab26b$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "trip_rate",
    color = NULL,
    title = "Average Trip Rates by Age"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Average Daily Trips by Ethnicity

```{r trip rate by Ethnicity}
tab28a <- Person %>%
  group_by(year, ethnicity_broad) %>%
  summarise(
    people = sum(person_weight),
  ) 

tab28b <- Trips %>%
  left_join(Person %>% select(c("person_id", "ethnicity_broad")), by = "person_id") %>%
  group_by(year, ethnicity_broad) %>%
  summarise(
    tot_per_trip = sum(trip_weight))

tab28 <- tab28b %>%
  left_join(tab28a, by = c("year", "ethnicity_broad")) %>%
  mutate(
    trip_rate = tot_per_trip/people
  )%>%
  select(-tot_per_trip, -people) %>%
  filter(ethnicity_broad != "")


# Create the line graph
ggplot(tab28, aes(x = year, y = trip_rate, color = ethnicity_broad, group = ethnicity_broad)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab26b$year), max(tab26b$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "trip_rate",
    color = NULL,
    title = "Average Trip Rates by Ethnicity"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```


------------------------------------------------------------------------

## Average Daily Trips by University Student Status

```{r trip rate by univ stud}
Person <- Person %>%
  mutate(univ_status = case_when(
    school_type >= 10 & school_type <= 13 ~ "University Student", 
    TRUE ~ "Not University Student"))

tab29a <- Person %>%
  group_by(year, univ_status) %>%
  summarise(
    people = sum(person_weight)
  ) 

tab29b <- Trips %>%
  left_join(Person %>% select(c("person_id", "univ_status")), by = "person_id") %>%
  group_by(year, univ_status) %>%
  summarise(
    tot_per_trip = sum(trip_weight))

tab29 <- tab29b %>%
  left_join(tab29a, by = c("year", "univ_status")) %>%
  mutate(
    trip_rate = tot_per_trip/people
  )%>%
  select(-tot_per_trip, -people)


# Create the line graph
ggplot(tab29, aes(x = year, y = trip_rate, color = univ_status, group = univ_status)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab26b$year), max(tab26b$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "trip_rate",
    color = NULL,
    title = "Average Trip Rates by University Student"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```


------------------------------------------------------------------------

## Average Daily Trips by Worker Status

```{r trip rate by worker status}
Person <- Person %>%
  mutate(worker_status = ifelse(is.na(num_jobs), "Not Employed", "Employed"))

tab30a <- Person %>%
  group_by(year, worker_status) %>%
  summarise(
    people = sum(person_weight)
  ) 

tab30b <- Trips %>%
  left_join(Person %>% select(c("person_id", "worker_status")), by = "person_id") %>%
  group_by(year, worker_status) %>%
  summarise(
    tot_per_trip = sum(trip_weight))

tab30 <- tab30b %>%
  left_join(tab30a, by = c("year", "worker_status")) %>%
  mutate(
    trip_rate = tot_per_trip/people
  )%>%
  select(-tot_per_trip, -people) 


# Create the line graph
ggplot(tab30, aes(x = year, y = trip_rate, color = worker_status, group = worker_status)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab26b$year), max(tab26b$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "trip_rate",
    color = NULL,
    title = "Average Trip Rates by Worker Status"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```


------------------------------------------------------------------------

## Travel for Work

Post pandemic, the share of auto trips for work travel has returned to pre-pandemic levels. The region’s very large decrease in bus ridership has continued post-pandemic. The decrease in bus ridership is likely because of a reduction in overall trips, as well as a reduction in bus service. 2022 had transit trips, but at a very low rate.

```{r work mode share}
tab17 <- Trips %>%
  filter(trip_purpose == "Home to Work" & mode_simple != "") %>%
  group_by(year, mode_simple) %>%
  summarise(
    trip = sum(trip_weight)) %>%
  mutate(
    Percentage = trip/sum(trip)) %>%
  select(-trip)

tab17$mode_simple <- factor(
  tab17$mode_simple,
  levels = c("auto", "bus", "bike", "walk", "school bus", "other")
)
# Create the line graph
ggplot(tab17, aes(x = year, y = Percentage, color = mode_simple, group = mode_simple)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab17$year), max(tab17$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    color = NULL,
    title = "Mode Share Across Years"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Travel for Shopping

Post pandemic, the share of auto trips for work travel has returned to pre-pandemic levels. The region’s very large decrease in bus ridership has continued post-pandemic. The decrease in bus ridership is likely because of a reduction in overall trips, as well as a reduction in bus service. 2022 had transit trips, but at a very low rate.

```{r shop mode share}
tab18 <- Trips %>%
  filter(trip_purpose == "Home to Shop" & mode_simple != "") %>%
  group_by(year, mode_simple) %>%
  summarise(
    trip = sum(trip_weight)) %>%
  mutate(
    Percentage = trip/sum(trip)) %>%
  select(-trip)

tab18$mode_simple <- factor(
  tab18$mode_simple,
  levels = c("auto", "bus", "bike", "walk", "school bus", "other")
)
# Create the line graph
ggplot(tab18, aes(x = year, y = Percentage, color = mode_simple, group = mode_simple)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab18$year), max(tab18$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    color = NULL,
    title = "Mode Share Across Years"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Travel for School

Auto and school bus trips are more similar to pre-pandemic levels. Proportionately, kids are biking and walking more to school.

```{r sch mode share}
tab18 <- Trips %>%
  filter(trip_purpose == "Home to School" & mode_simple != "") %>%
  group_by(year, mode_simple) %>%
  summarise(
    trip = sum(trip_weight)) %>%
  mutate(
    Percentage = trip/sum(trip)) %>%
  select(-trip)

tab18$mode_simple <- factor(
  tab18$mode_simple,
  levels = c("auto", "bus", "bike", "walk", "school bus", "other")
)
# Create the line graph
ggplot(tab18, aes(x = year, y = Percentage, color = mode_simple, group = mode_simple)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab18$year), max(tab18$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    color = NULL,
    title = "Mode Share Across Years"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Travel for University

Although the shares of university travel by mode have shifted fairly significantly, these trips still hold a majority with auto followed by non-motorized modes. As seen for other trip purposes, transit shares have reduced post-pandemic.  

```{r univ mode share}
tab19 <- Trips %>%
  filter(trip_purpose == "Home to University") %>%
  group_by(year, mode_simple) %>%
  summarise(
    trip = sum(trip_weight)) %>%
  mutate(
    Percentage = trip/sum(trip)) %>%
  select(-trip)

tab19$mode_simple <- factor(
  tab19$mode_simple,
  levels = c("auto", "bus", "bike", "walk", "school bus", "other")
)
# Create the line graph
ggplot(tab19, aes(x = year, y = Percentage, color = mode_simple, group = mode_simple)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab19$year), max(tab19$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    color = NULL,
    title = "Mode Share Across Years"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Travel for Other Reasons – Home to Other

At close to 15% of the overall mode share, walking trips are most competitive for trips made from home for activities like recreation, social, religious, dining and medical.


```{r hbo mode share}
tab20 <- Trips %>%
  filter(trip_purpose == "Home to Other" & mode_simple != "") %>%
  group_by(year, mode_simple) %>%
  summarise(
    trip = sum(trip_weight)) %>%
  mutate(
    Percentage = trip/sum(trip)) %>%
  select(-trip)

tab20$mode_simple <- factor(
  tab20$mode_simple,
  levels = c("auto", "bus", "bike", "walk", "school bus", "other")
)
# Create the line graph
ggplot(tab20, aes(x = year, y = Percentage, color = mode_simple, group = mode_simple)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab20$year), max(tab20$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    color = NULL,
    title = "Mode Share Across Years"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Travel for Other Reasons – Other to Other

At close to 15% of the overall mode share, walking trips are most competitive for trips made from home for activities like recreation, social, religious, dining and medical.


```{r nhb mode share}
tab21 <- Trips %>%
  filter(trip_purpose == "Other to other" & mode_simple != "") %>%
  group_by(year, mode_simple) %>%
  summarise(
    trip = sum(trip_weight)) %>%
  mutate(
    Percentage = trip/sum(trip)) %>%
  select(-trip)

tab21$mode_simple <- factor(
  tab21$mode_simple,
  levels = c("auto", "bus", "bike", "walk", "school bus", "other")
)
# Create the line graph
ggplot(tab21, aes(x = year, y = Percentage, color = mode_simple, group = mode_simple)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab21$year), max(tab21$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    color = NULL,
    title = "Mode Share Across Years"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Other Travel Metrics

This section focuses on special travel patterns – non-motorized trips, travel by seniors, work related travel, and travel post pandemic. 


------------------------------------------------------------------------

## Non-motorized Trips

Non-motorized travel for non-recreational trips is slowly increasing back from the decline during the pandemic. Trips under 1 mile remain high and longer distance trips increased significantly, perhaps due to the new e-mobility options available on the market.

:::: {.columns}
::: {.column width="50%"}
```{r nm mode share}
tab22a <- Trips %>%
  filter(mode_simple == "walk" | mode_simple == "bike") %>%
  group_by(year) %>%
  summarise(
    nmtrips = sum(trip_weight, na.rm = T)
  )

tab22b<- Trips %>%
  group_by(year) %>%
  summarise(
    trips = sum(trip_weight, na.rm = T)
  )

tab22 <- merge(tab22a, tab22b, by = "year") %>%
  mutate(
    Percentage = nmtrips/trips) %>%
  select(-nmtrips, -trips)

tab22_t <- tab22 %>%
  tibble::column_to_rownames("year") %>%
  t()

tab22_t %>%
  as.data.frame() %>%
  tibble::rownames_to_column(" ") %>%
  gt() %>%
  fmt_percent(columns = -` `, decimals = 0)

```

:::
::: {.column width="50%"}
```{r nm distance}
tab22c <- Trips %>%
  filter(mode_simple == "walk" | mode_simple == "bike") %>%
  group_by(year, distbin) %>%
  summarise(
    nmtrips = sum(trip_weight, na.rm = T)
  ) %>%
  mutate(
    Percentage = nmtrips/sum(nmtrips)) %>%
  select(-nmtrips)

tab22c$distbin <- factor(
  tab22c$distbin,
  levels = c("Under 1 Mile", "Between 1 - 6 Miles", "Over 6 Miles")
)
# Create the line graph
ggplot(tab22c, aes(x = year, y = Percentage, color = distbin, group = distbin)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab22c$year), max(tab22c$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    color = NULL,
    title = "Distance Traveled"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```
::::


------------------------------------------------------------------------

## Non-motorized Trips

Non-motorized travel for non-recreational trips is slowly increasing back from the decline during the pandemic. Trips under 1 mile remain high and longer distance trips increased significantly, perhaps due to the new e-mobility options available on the market.

```{r nm purpose}
tab22d <- Trips %>%
  filter(mode_simple == "walk" | mode_simple == "bike") %>%
  filter(!is.na(trip_purpose))%>%
  group_by(year, trip_purpose) %>%
  summarise(
    nmtrips = sum(trip_weight, na.rm = T)
  ) %>%
  mutate(
    Percentage = nmtrips/sum(nmtrips)) %>%
  select(-nmtrips)

tab22d$trip_purpose <- factor(
  tab22d$trip_purpose,
  levels = c("Home to Work","Home to Shop","Home to Other","Home to University","Home to School","Other to other")
)

# Create the stacked area graph
ggplot(tab22d, aes(x = year, y = Percentage, fill = trip_purpose)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9) +
  scale_x_continuous(
    breaks = seq(min(tab22d$year), max(tab22d$year), 1),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Non-Motorized Traveler Trip Purpose"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```


------------------------------------------------------------------------

## Travel by Seniors

:::: {.columns}
::: {.column width="50%"}
```{r senior mode share}
tab23a <- Trips %>%
  left_join(Person %>% select(c("person_id", "agecat")), by="person_id") %>%
  filter(agecat == "65+" & mode_simple!="") %>%
  group_by(year, mode_simple) %>%
  summarise(
    trip = sum(trip_weight, na.rm = T)
  )%>%
  mutate(
    Percentage = trip/sum(trip)) %>%
  select(-trip)%>%
  pivot_wider(names_from = "mode_simple", values_from = 'Percentage') %>%
  select(c("year", "auto", "bus", "bike", "walk", "other"))

tab23a %>%
  as.data.frame() %>%
  gt() %>%
  fmt_percent(columns = any_of(c("auto", "bus", "bike", "walk", "other")), decimals = 0)

```

:::
::: {.column width="50%"}
```{r senior trip purp}
tab23b <- Trips %>%
  left_join(Person %>% select(c("person_id", "agecat")), by="person_id") %>%
  filter(agecat == "65+") %>%
  group_by(year, trip_purpose) %>%
  summarise(
    trip = sum(trip_weight, na.rm = T)
  )%>%
  mutate(
    Percentage = trip/sum(trip)
  ) %>%
  select(-trip) %>%
  filter(trip_purpose!= "Home to University")

tab23b$trip_purpose <- factor(
  tab23b$trip_purpose,
  levels = c("Home to Work","Home to Shop","Home to Other","Other to other"))


# Create the line graph
ggplot(tab23b, aes(x = year, y = Percentage, color = trip_purpose, group = trip_purpose)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab23b$year), max(tab23b$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    color = NULL,
    title = "Reason for Travel by Seniors"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```
::::

------------------------------------------------------------------------

## Work Related - Telework Patterns

:::: {.columns}
::: {.column width="50%"}
```{r work location}
tab24a <- Person %>%
  filter(job_type!="") %>%
  mutate(
    job_type_simple = ifelse(
      job_type == "Go to one work location outside the home", "Go to one work location outside the home (may also telework)", job_type
  )) %>%
  group_by(year, job_type_simple) %>%
  summarise(
    people = sum(person_weight),
  ) %>%
  mutate(
    Percentage = people/sum(people)
  ) %>%
  select(-people)%>%
  mutate(
    year = factor(year, levels = sort(unique(year)))
  )

ggplot(tab24a, aes(x = job_type_simple, y = Percentage, fill = year)) +
  geom_col(
    position = position_dodge2(width = 0.8, preserve = "single"),
    width = 0.7
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_wolfpack() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 16)) +
  labs(
    x = NULL, y = "Percentage", fill = "Year",
    title = "Job Type Distribution by Year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, lineheight = 0.95)
  )

```
:::
::: {.column width="50%"}
```{r telework freq}
tab24b <- Person %>%
  filter(!is.na(telework_freq)) %>%
  mutate(
    year = as.character(year),
    telework_freq_simple = case_when(
      telework_freq == "4 days a week" | telework_freq == "2-3 days a week" ~ "2-4 days a week",
      telework_freq == "1 day a week" | telework_freq == "1 day every 2 weeks" | telework_freq == "1-3 days per month" ~ "1-4 days a month",
      TRUE ~ telework_freq
    )) %>%
  group_by(year, telework_freq_simple) %>%
  summarise(
    people = sum(person_weight)
  )%>%
  mutate(
    Percentage = people/sum(people)
  ) %>%
  select(-people)

ggplot(tab24b, aes(x = telework_freq_simple, y = Percentage, fill = year)) +
  geom_col(
    position = position_dodge2(width = 0.8, preserve = "single"),
    width = 0.7
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_wolfpack() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 16)) +
  labs(
    x = NULL, y = "Percentage", fill = "Year",
    title = "Telework Frequency by Year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, lineheight = 0.95)
  )
```
:::
::::

------------------------------------------------------------------------

## Time of Day


```{r tod}
tab31 <- Trips %>%
  group_by(year, tod) %>%
  summarise(
    trip = sum(trip_weight)
  ) %>%
  mutate(
    Percentage = trip/sum(trip)
  ) %>%
  select(-trip)

tab31$tod <- factor(
  tab31$tod,
  levels = c("AM", "PM", "MD", "NT"))

# Stacked area chart (Time of Day pattern)
ggplot(tab31, aes(x = year, y = Percentage, fill = tod)) +
  geom_area(color = "white", size = 0.2, alpha = 0.9) +
  scale_x_continuous(
    breaks = seq(min(tab31$year), max(tab31$year), 1),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    fill = NULL,
    title = "Time of Day Pattern"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

------------------------------------------------------------------------

## Work Related - Departure Time and Trip Length

:::: {.columns}
::: {.column width="50%"}
```{r Departure Time}
tab25a <- Trips %>%
  mutate(
    starttime = floor(departure_time/60),
    starttime = ifelse(starttime>24, starttime-24, starttime),
    year = as.character(year)
  ) %>%
  group_by(year, starttime) %>%
  summarise(
    trip = sum(trip_weight)
  ) %>%
  mutate(
    Percentage = trip/sum(trip)
  ) %>%
  select(-trip)

# Create the line graph
ggplot(tab25a, aes(x = starttime, y = Percentage, color = year, group = year)) +
  geom_line(size = 1) +
  scale_x_continuous(limits = c(0, 24), breaks = 0:24, expand = c(0, 0)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Percentage",
    color = NULL,
    title = "Departure Time of Commute Trips"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```
:::
::: {.column width="50%"}
```{r Commute trip length}
tab25b <- Trips %>%
  filter(trip_purpose == "Home to Work" & mode_simple!= ""& mode_simple!= "other") %>%
  group_by(year, mode_simple) %>%
  summarise(
    tlength = weighted.mean(distance_miles, trip_weight, na.rm = T)
  ) 

tab25b$mode_simple <- factor(
  tab25b$mode_simple,
  levels = c("auto", "bus", "bike", "walk")
)

# Create the line graph
ggplot(tab25b, aes(x = year, y = tlength, color = mode_simple, group = mode_simple)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(tab25b$year), max(tab25b$year), 1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  scale_color_wolfpack() +
  labs(
    x = "Year",
    y = "Trip Length",
    color = NULL,
    title = "Average Commute Trip Length"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

:::
::::
