---
title: "survey_combined"
author: "Kyle, Si"
date: "9/14/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(dplyr.summarise.inform = FALSE)
options(scipen = 999)

library(tidyverse)
library(broom)
library(readxl)
library(ggplot2)
library(plotly)
library(leaflet)
library(sf)
library(geosphere)
library(knitr)
library(scales)
library(kableExtra)
library(openxlsx)
library(XLConnect)
library(sp)
```


```{r read_data, message=FALSE}
setwd("C:/Work/Travel Trends/HTS/2024 HTS/ITRE_combined_weighted_deliverables_2025-07-01")
# Read Data

## Household files
hh_raw <- read_csv("hh.csv")

## Person files
per_raw <- read_csv("person.csv") 

## Trip files
trips_raw <- read_csv("trip.csv") 

## Veh files
veh_raw <- read_csv("vehicle.csv")

## Day file(s)
day_raw <- read_csv("day.csv")

## Conversion files
hhconvfile <- "C:/Git/TravelTrendAnalysis/Equiv/2024/HH_data_conversion.xlsx"
perconvfile <- "C:/Git/TravelTrendAnalysis/Equiv/2024/Person_data_conversion.xlsx"
vehconvfile <- "C:/Git/TravelTrendAnalysis/Equiv/2024/Vehicle_data_conversion.xlsx"
tripconvfile <- "C:/Git/TravelTrendAnalysis/Equiv/2024/Trip_data_conversion.xlsx"
dayconvfile <- "C:/Git/TravelTrendAnalysis/Equiv/2024/Day_data_conversion.xlsx"

```

## Introduction

The purpose of this script is to combine 2016, 2018, and 2021 household travel
survey into one data set and then analyze travel behavior.

## Survey combination

2016 and 2018 survey look similar while 2021 is a different animal. In 2021, 
r_Move phone app was introduced so participants are able to fill out a 7-day 
travel diary instead of one. Also, the questions asked changes drastically in
2021. Hence, a crosswalk table is developed in order to combine the data set. In
the following section, each survey will be processed independently based on the 
crosswalk and then merge into one.

### Household table

```{r household}
## Household Table

remove_na_hh <- hh_raw %>%
  filter(hh_weight>0) %>%
#  rename(
#    hhweight = hh_weight,
#    residence_type = res_type,
#    income_aggregate = income_broad,
#    num_bicycles = num_bicycle,
#    num_children = num_kids,
#    num_surveyable_people = num_participants
#  ) %>%
  mutate(
    hh_id = paste0(year,"_", hh_id)) %>%
  select(-c(
    sample_segment, signup_platform, num_complete_mon:num_complete_sun,
    participate_future, hh_income_imputed, first_travel_date, last_travel_date,
    num_days_complete_weekday, num_days_complete_weekend)
  )

### Join Equiv/crosswalk table
hhconv <- XLConnect::loadWorkbook(hhconvfile)
sheets <- XLConnect::getSheets(hhconv)

for (tab in sheets){
  colnames(remove_na_hh)[colnames(remove_na_hh) == tab] <- "Value"
  conv <- read_excel(hhconvfile, sheet = tab)
  remove_na_hh <- left_join(remove_na_hh, conv, 
                         by=c("Value" = "Value", "year" = "Year")) %>%
    select(-Value, -Translation)
  colnames(remove_na_hh)[colnames(remove_na_hh) == "CommLabel"] <- tab 
}

rm(hhconv, sheets, tab)

remove_na_hh <- remove_na_hh %>% 
  mutate(
    HHSize = ifelse(num_people < 4, as.character(num_people), "4+"),
    HHWorkers = ifelse(num_workers < 3, as.character(num_workers), "3+"),
    nveh = ifelse(num_vehicles < 3, as.character(num_vehicles), "3+"),
    vehratio = num_vehicles/num_workers,
    autoown = case_when(
      num_vehicles == 0 ~ "No Car",
      vehratio < 1 ~ "Less than One Car per Worker",
      TRUE ~ "One or More Cars per Worker"
    ))
```


### Person table

The person table changed substantially between 2016, 2018, and 2021. As an example,
race included 6 categories in 2016 and was contained in a single column. In
2018, race only had 5 categories and was contained in 5 separate binary columns
(e.g. `race_white` = 1 or 0). These formats had to be rectified, and the "Two+"
race category from 2016 collapsed into "Other", before the tables could be
combined.

In addition, some fields in the 2016/2018 tables were not present in 2021 (and vice
versa). The goal is to preserve as many data fields as I can while dropping fields 
irrelevant to travel behavior.


```{r person}
remove_na_per <- per_raw %>%
  filter(person_weight>0) %>%
  mutate(
    person_id = paste0(year,"_", person_id),
    hh_id = paste0(year,"_", hh_id)) %>%
  left_join(
    remove_na_hh %>% select(hh_id, hh_weight),
    by = "hh_id"
  ) %>%
  filter(hh_weight>0) %>%
  select(-c(ev_typical_charge_1:ev_typical_charge_997, work_post_covid_puma))

### Collapse race and ethnicity
# No longer need to do so because RSG created a race_broad/ethnicity_broad field
remove_na_per <- remove_na_per %>%
  select(-c(race_1:race_other, ethnicity_1:ethnicity_other))

### Join Per Equiv/crosswalk table
perconv <- XLConnect::loadWorkbook(perconvfile)
sheets <- XLConnect::getSheets(perconv)

for (tab in sheets){
  if(tab %in% colnames(remove_na_per)){
    colnames(remove_na_per)[colnames(remove_na_per) == tab] <- "Value"
    remove_na_per <- left_join(remove_na_per, read_excel(perconvfile, sheet = tab), 
                               by=c("year" = "Year", "Value" = "Value")) %>%
      select(-Value, -Translation)
    colnames(remove_na_per)[colnames(remove_na_per) == "CommLabel"] <- tab
  }
}

rm(perconv, sheets, tab)

```


### Vehicle table
```{r vehicle}
# prep_veh_2016 <- veh_2016_raw %>%
# Add if necessary  



```

### Day table
Unlike previous year, 2021 HTS is a multi-day survey. So household/person should 
be repeated for each day they travel. The sum of the day_weight in the day table 
equals the number of persons in the survey region. Day weights are not provided 
for 2016 and 2018 data because they are identical to the person weights(which is also household weight).
```{r day}
remove_na_day <- day_raw %>%
    filter(day_weight>0)%>%
    mutate(
      day_id = paste0(year,"_", day_id),
      person_id = paste0(year,"_", person_id),
      hh_id = paste0(year,"_", hh_id)) %>%
    left_join(
      remove_na_per %>% select(hh_id, person_id, hh_weight),
      by = c("hh_id", "person_id")
    ) %>%
    filter(hh_weight>0)


### Join Day Equiv table
dayconv <- XLConnect::loadWorkbook(dayconvfile)
sheets <- XLConnect::getSheets(dayconv)

for (tab in sheets){
  if(tab %in% colnames(remove_na_day)){
    colnames(remove_na_day)[colnames(remove_na_day) == tab] <- "Value"
    remove_na_day <- left_join(remove_na_day, read_excel(dayconvfile, sheet = tab), 
                               by=c("year" = "Year", "Value" = "Value")) %>%
      select(-Value, -Translation)
    colnames(remove_na_day)[colnames(remove_na_day) == "CommLabel"] <- tab
  }
}

rm(dayconv, sheets, tab)
```

### Trip table
```{r trip}
remove_na_trip <- trips_raw %>%
  mutate(
    trip_id = paste0(year,"_", trip_id),
    day_id = paste0(year,"_", day_id),
    person_id = paste0(year,"_", person_id),
    hh_id = paste0(year,"_", hh_id),
    travel_date = as.character(travel_date), ##2022 travel_date is in Date format originally
    travel_date = as.POSIXct(travel_date, format = "%Y-%m-%d"),
    departure_string = paste0(depart_date, " ",
                              str_pad(depart_hour, 2, pad = "0"), ":", 
                              str_pad(depart_minute, 2, pad = "0"), ":", 
                              str_pad(depart_second, 2, pad = "0")),
    departure_string = as.POSIXct(departure_string, format = "%Y-%m-%d %H:%M:%S"),
    arrival_string = paste0(arrive_date, " ",
                           str_pad(arrive_hour, 2, pad = "0"), ":", 
                           str_pad(arrive_minute, 2, pad = "0"), ":", 
                           str_pad(arrive_second, 2, pad = "0")),
    arrival_string = as.POSIXct(arrival_string, format = "%Y-%m-%d %H:%M:%S"),
    dtm = paste0(travel_date, " ", "00:00:00"),
    dtm = as.POSIXct(dtm, format = "%Y-%m-%d %H:%M:%S"),
    departure_time = as.numeric(difftime(departure_string, dtm, units = "mins")),
    arrival_time = as.numeric(difftime(arrival_string, dtm, units = "mins")),
#    vehiclenum = case_when(
#      mode_1 >=6 & mode_1<= 12 ~ mode_1
#    )
    ) %>%
  select(-c(mode_1:mode_4, o_puma_2012:o_taz, d_puma_2012:d_taz, depart_dow:arrive_second, dtm, 
            ev_charge_station:ev_charge_station_level_998,
            park_cost, taxi_cost, o_purpose_other, d_purpose_other, o_activity_derived:d_activity_itre)) %>%
  filter(trip_weight>0) %>%
    left_join(
      remove_na_hh %>% select(hh_id, hh_weight),
      by = "hh_id"
    ) %>%
    filter(hh_weight>0)

### Join Trip Equiv table
tripconv <- XLConnect::loadWorkbook(tripconvfile)
sheets <- XLConnect::getSheets(tripconv)

for (tab in sheets){
  if(tab %in% colnames(remove_na_trip)){
    colnames(remove_na_trip)[colnames(remove_na_trip) == tab] <- "Value"
    remove_na_trip <- left_join(remove_na_trip, read_excel(tripconvfile, sheet = tab), 
                               by=c("year" = "Year", "Value" = "Value")) %>%
      select(-Value, -Translation)
    colnames(remove_na_trip)[colnames(remove_na_trip) == "CommLabel"] <- tab
  }
}

rm(tripconv, sheets, tab)

```

```{r check missing o activity}

### Check for missing O activities if it is first trip in a day
num_missing <- remove_na_trip %>%
  filter(is.na(o_purpose)) %>%
  group_by(year) %>%
  summarise(record_count = n())
   #54 records in 2021 and 1629 records in 2022 and 1456 records in 2024 do not have o_purpose

checkmissing <- remove_na_trip %>%
  left_join(remove_na_day %>% select(day_id, begin_day), 
  by = "day_id") %>%
  mutate(
    o_purpose = case_when(
      is.na(o_purpose) & trip_num == 1 ~ begin_day,  #all records are first trip in a day
      is.na(o_purpose) ~ "",
      TRUE ~ o_purpose
    )
  )

num_missing_update <- checkmissing %>%
  filter(is.na(o_purpose)) %>%
  group_by(year) %>%
  summarise(record_count = n()) ## still 5 records from 2022 and 9 records from 2024 have missing o_purpose because begin_day is also null. O_lat/lon is available so maybe able to determine later
  
```

## Basic checks

A common task in household survey processing is to remove trips/households for various
reasons. For instance, travel models estimate weekday travel, so households
surveyed on the weekend must be removed. Households may have incomplete travel
diaries, be located outside the model region, or have other issues that means
they must be removed from estimation set.

### Day of week

Thanks to careful survey design, none of the Triangle households were surveyed on the weekend. The chart below shows the days surveyed.

```{r}
temp <- checkmissing %>%
  group_by(year, travel_dow) %>%
  summarize(samples = n())

ggplot(temp, aes(x = travel_dow, y = samples, fill = as.factor(travel_dow))) +
  geom_bar(stat = "identity") +
  guides(fill = "none") +
  facet_wrap(~year)

rm(temp)
```

### Household and trip locations 

For travel trend analysis, it is important to make sure we are only focusing 
on a uniform geography where people shares similar travel behavior. In this case, 
we limit our analysis area down to model boundary. Any household outside model
boundary will be removed.

```{r hh_locations}
if(1==0){
hh_shp <- remove_na_hh %>%
  select(home_lat, home_lon) %>%
  st_as_sf(
    coords = c("home_lon", "home_lat"), crs = 4326, agr = "constant"
  )

boundary_shp <- st_read(
  "C:/Work/Travel Trends/HTS/Boundary/boundary_v2.shp") #boundary_shp seems empty

boundary_shp <- st_transform(boundary_shp, st_crs(shp))
bbox <- as.list(st_bbox(boundary_shp))
lng <- (bbox$xmin - bbox$xmax) / 2 + bbox$xmax
lat <- (bbox$ymax - bbox$ymin) / 2 + bbox$ymin

leaflet(
  options = leafletOptions(zoomControl = FALSE, minZoom = 9, maxZoom = 9)
) %>%
  addTiles() %>%
  addPolygons(
    data = boundary_shp,
    color = "#444444", weight = 1, smoothFactor = 0.5
  ) %>%
  addCircleMarkers(
    data = hh_shp, radius = 1, weight = 3,
    opacity = .2, fillOpacity = .2
  ) %>%
  setView(lng, lat, zoom = 9)

}
```

Similar to removing household records that are not in the triangle region, trip 
records where one end falls outside of model boundary should also be removed. By
filtering out these records, we are only focusing on the internal-internal trips. In 2024 deliverables, o_in_region and d_in_region are already provided.

```{r trip_locations }

otrip_shp <- checkmissing %>%
  select(o_lat, o_lon) %>%
  filter(!is.na(o_lat)&!is.na(o_lon)) %>%
  st_as_sf(
    coords = c("o_lat", "o_lon"), crs = 4326, agr = "constant"
  )

dtrip_shp <- checkmissing %>%
  select(d_lat, d_lon) %>%
  filter(!is.na(d_lat)&!is.na(d_lon)) %>%
  st_as_sf(
    coords = c("d_lat", "d_lon"), crs = 4326, agr = "constant"
  )

IItrip <- checkmissing %>%
  filter(o_in_region == 1 & d_in_region ==1)
```

### Weight distribution

It is important to check the weighting and expansion of a survey for extreme
weight values. To do this, the weights must first be normalized based on the
expected average weight per sample. A simple comparison of two hypothetical
surveys explains why:

1. A survey with 100 samples of a population of 100,000: Average weight = 1,000
2. A survey with 1000 samples of a population of 100,000: Average weight = 100

A household weight of 1,000 in the first survey is reasonable based on the
number of samples, but it is 10 times the expected weight in the second survey.
For the combined survey, the average weight is shown below.

```{r}
temp <- remove_na_hh %>%
  summarize(
    Samples = n(),
    `Total Households` = sum(hh_weight),
    `Average Weight` = `Total Households` / Samples
  ) %>%
  pivot_longer(cols = everything())

weight_mean <- mean(remove_na_hh$hh_weight)
colnames(temp) <- c("", "")

temp %>%
  kable(digits = 0) %>%
  kable_styling(full_width = FALSE)
```

The first histogram below shows the distribution of weights divided by the mean
weight. While the largest weight is 26x the mean weight, this is an outlier and
most are within 6x. This is a reasonable result and indicates and well-designed
sample plan that collected adequate numbers of households in all the demographic
groups of interest.

```{r}
ggplot(remove_na_hh, aes(x=hh_weight / weight_mean)) + 
  geom_histogram(color="black", fill="orange", bins = 20) +
  xlab("Weight / Average Weight")
```

The second histogram below inverts the factor above and divides the average
weight by the individual households weights. Large factors in this histogram
would indicate the presence of small weights. The smallest weight (11.6) is 4%
of the average weight, suggesting a modest over sample, but is well within
reason.

```{r}
ggplot(remove_na_hh, aes(x= weight_mean/ hh_weight)) + 
  geom_histogram(color="black", fill="orange", bins = 20) +
  xlab("Average Weight / Weight")
```

After a review of RSGs expansion procedure and the review of weights above,
Caliper determined that the weights were appropriate for use in model
estimation.

### Income imputation
RSG stops providing imputed income in 2021. So this section is left blank.


## Trip processing

### Activity purpose
he atomic model purposes are shown below. These represent the fundamental
activity purposes and may be aggregated during the estimation of specific model
components.

- H: Home (non-work activities at home)
- W: Work
- WR: Work-related (e.g. making a delivery)
- SCH: School. Clarified explicitly in the person table into:
  - PreK: Pre-Kindergarten
  - K12: Kindergarten through 12th grade
  - U: University or College
- SHP: Shop
- EAT: Dining out
- PU: Pick up
- DO: Drop off
- X: Transfer modes (e.g. at a bus stop)
- OM: Other maintenance 
- OMED: Other medical
- OD: Other discretionary

The table below shows how the survey purposes (for all years) were collapsed
into the purposes above.
```{r}
temp <- read_excel(tripconvfile, sheet = "o_purpose") 
temp %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped")
```

```{r OD purpose}
convert_purpose <- IItrip %>%
  left_join(
    remove_na_per %>% select(person_id, school_type, age),
    by = "person_id"
  ) %>%
  mutate(
    # some people marked 1-yo old as student with school at home. Not what
    # we are interested in.
    student_type_broad = case_when(
      school_type >= 4 & school_type <= 9 ~ "K12",
      school_type >= 10 & school_type <= 13 ~ "U",
      school_type==2 | school_type==3 ~ "PreK",
      TRUE ~ ""
    ),
    o_purpose = case_when(
      o_purpose != "SCH" ~ o_purpose,
      student_type_broad != "" ~ student_type_broad,
      TRUE ~ "SCH?"
    ),
    d_purpose = case_when(
      d_purpose != "SCH" ~ d_purpose,
      student_type_broad != "" ~ student_type_broad,
      TRUE ~ "SCH?"
    ),
    homeschooled_flag = ifelse(school_type %in% c(5, 7, 9), 1, 0)
  )

```

### Places

As explained above, the definition of a work activity in the 2016 survey did not
specify whether work was performed in or outside the home. Caliper computed a
place code for every trip end independent of activity.

- H: Home
- W: Work
- S: School
- O: Other

Home location was derived by comparing the activity location (long/lat) with 
the home locations. Trip ends within 100 meters of the home were tagged to the 
appropriate place type. This approach gave a more accurate result than depending 
on the names each person gave their origin and destination, which varied widely. 
It also handled several cases where respondents separated places like "front 
yard" and "shed" from their physical home.

School location was handled differently for 2016/2018 and 2021. For 2016/2018
data, school address and school name were available and were used to determine if 
the destination is school. However, for 2021 survey data, address and place names
were missing so distance is used to impute place type.

Work locations could not be handled the same way. Not only do people have
multiple jobs, but many had jobs without fixed work locations (e.g. nanny). For
this reason, we relied on the stated activity purpose to determine work places.

```{r determine_place PleaseReview}
# Use calculated distances (rather than place names) to determine trip end
# place type (H, W, S(chool), O)
calc_dist <- function(x1, y1, x2, y2) {
  df1 <- tibble(lng = x1, lat = y1)
  df2 <- tibble(lng = x2, lat = y2)
  distHaversine(df1, df2)
}

det_home_place <- convert_purpose %>%
  # Home
  left_join(
    remove_na_hh %>% select(hh_id, home_lat, home_lon),
    by = "hh_id"
  ) %>%
  mutate(
    o_home_dist = calc_dist(
      o_lon, o_lat, home_lon, home_lat),
    d_home_dist = calc_dist(
      d_lon, d_lat, home_lon, home_lat),
    o_place = ifelse(o_home_dist < 30, "H", "O"), 
    o_place = ifelse(is.na(o_place), "O", o_place),
    d_place = ifelse(d_home_dist < 30, "H", "O"),
    d_place = ifelse(is.na(d_place), "O", d_place),
  ) %>%
  select(-home_lat, -home_lon) %>%
  # correct home activities that don't happen at home
  mutate(
    o_purpose = ifelse(o_purpose == "H" & o_place != "H", "O", o_purpose),
    d_purpose = ifelse(d_purpose == "H" & d_place != "H", "O", d_purpose)
  )
  # there are people mark d_purpose as "X" while d_place == H. this typically happens where people return home with car and then ride bike for some other purposes.

school_loc <- remove_na_per %>%
  select(hh_id, person_id, year, school_lat, school_lon) %>%
  left_join(
    remove_na_hh %>% select(hh_id, home_lat, home_lon),
    by = "hh_id"
  ) %>%
  mutate(hsdist = calc_dist(home_lon, home_lat, school_lon, school_lat)) %>%
  filter(!is.na(school_lat)& hsdist>100) %>% 
  select(-hh_id, -home_lat, -home_lon, -hsdist, -year)


# only check S place for K-12 and prek students
det_sch_place <- det_home_place %>%
  left_join(school_loc, by = "person_id") %>%
  mutate(
    o_sch_dist = calc_dist(o_lon, o_lat, school_lon, school_lat),
    d_sch_dist = calc_dist(d_lon, d_lat, school_lon, school_lat),
    o_place = case_when(
      is.na(o_sch_dist) ~ o_place,
      o_place == "H" ~ "H",
      o_sch_dist < 600 ~ "S",
      # here we use a broad radius for school.
      TRUE ~ o_place
    ),
    d_place = case_when(
      is.na(d_sch_dist) ~ d_place,
      d_place == "H" ~ "H",
      d_sch_dist < 600 ~ "S",
      #here we use a broad radius for school.
      TRUE ~ d_place
    ),
    # for students with missing school loc info, rely on reported activity
    o_place = ifelse(
      student_type_broad != "" &
        is.na(school_lon) &
        o_purpose %in% c("PreK", "K12", "U") &
        o_place == "O", 
      "S", o_place
    ),
    # for students with missing school loc info, rely on reported activity
    d_place = ifelse(
      student_type_broad != "" &
        is.na(school_lon) &
        d_purpose %in% c("PreK", "K12", "U") &
        d_place == "O", 
      "S", d_place
    )
  ) %>%
  select(-school_lon, -school_lat)



# Determining work place is harder. People have multiple jobs, and some jobs
# don't have fixed addresses. Rely on their reported activity.
det_work_place <- det_sch_place %>%
  mutate(
    o_place = case_when(
      o_place == "H" ~ "H",
      o_purpose == "W" ~ "W",
      TRUE ~ o_place
    ),
    d_place = case_when(
      d_place == "H" ~ "H",
      d_purpose == "W" ~ "W",
      TRUE ~ d_place
    )
  ) %>%
  relocate(o_place, .before = o_purpose) %>%
  relocate(d_place, .after = o_place)
```

### Remove home-home loop trip
There are trips that start and end both at home. Remove those and change the start of next trip to home
```{r removeloop}
remove_loop <- det_work_place %>%
  group_by(person_id, day_num) %>%
  arrange(trip_num, .by_group = TRUE) %>%
  mutate(
    remove_flag = case_when(
      o_purpose == "H" & d_purpose == "H" ~ 1,
      o_purpose == "H" & d_purpose == "X" & lead(o_place) == "H" ~ 1,
      TRUE ~ 0 )) %>%
  # If current trip is home-home, set next trip's origin to home
  mutate(
    o_purpose = ifelse(lag(remove_flag, default = FALSE), "H", o_purpose)) %>%
  filter(remove_flag!=1) 
```


### Tour formation

This travel trend analysis will be done at tour level, instead of trip level. Since
the current HTS is at trip level, we need to develop a method to combine trips 
into tour.

For each person in the survey, a tour begins with their first trip and ends
when they return home. A second tour starts if they leave home again. A tour is
classified as a work tour if the traveler has a work activity during the tour.
If they do not, then the tour is classified as non-work.

There is a third tour type ("home"), that handles cases where the survey reports
trips within the home. For example, when a respondent reports being at home and
then working from home, the survey includes that as a trip. Home tours also
capture things like recreational walks that start and end at the home. For the
purpose of travel modeling, these home tours are ignored.

```{r tour_formation}
determine_tours <- remove_loop %>%
  mutate(
    new_tour = ifelse(row_number() == 1, 1, 0),
    new_tour = ifelse(o_place == "H" & lead(o_place) != "H" & lead(d_place) != "H", 1, new_tour),
    tour_num = cumsum(new_tour)
  ) %>%
  group_by(person_id, day_id, tour_num) %>%
  arrange(trip_num, .by_group = TRUE) %>%
  mutate(
    trips_in_tour = n(),
    tour_distance = sum(distance_miles),
    tour_duration = last(arrival_time) - first(departure_time), ##2016 and 2018 have records with negative departure time
    tour_anchor = ifelse(d_purpose == "W", "W", ""),
    tour_anchor = ifelse(o_place == "H" & d_place == "H", "H", tour_anchor),
    tour_type = case_when(
      "W" %in% tour_anchor ~ "WT",
      # "S" %in% tour_anchor ~ "S",
      "H" %in% tour_anchor ~ "H",
      TRUE ~ "NT"
    ),
    o_party_size_change = num_travelers - lag(num_travelers),
    d_party_size_change = lead(num_travelers) - num_travelers,
    o_act_dur = departure_time - lag(arrival_time),
    d_act_dur = lead(departure_time) - arrival_time
  )

# loop trips are things like recreational walks (loop tour here are coded as seperate trip records). For instance, home to pick up mail and then return home as two separate trips.
flag_loop_tours <- determine_tours %>%
  mutate(
    loop_flag = case_when(
      trips_in_tour > 2 ~ 0,
      d_act_dur > 10 ~ 0,
      !(mode_type %in% c("walk", "bike")) ~ 0,
      !(o_purpose %in% c("H", "OD")) ~ 0,
      !(d_purpose %in% c("H", "OD")) ~ 0,
      TRUE ~ 1
    )
  )

```


### Problems

```{r}
sch_problems <- flag_loop_tours %>%
  mutate(
    sch_flag = case_when(
      o_purpose == "SCH?" ~ 1,
      d_purpose == "SCH?" ~ 1,
      TRUE ~ 0
    )
  ) %>%
  filter(sch_flag == 1)

num_sch_problems <- sch_problems %>%
  group_by(year) %>%
  summarise(record_count = n())
  

#temp <- flag_loop_trips %>%
#  ungroup() %>%
#  filter(linked_trip_id %in% num_sch_problems)
  
  
##812, 298, 32, 77, 27 records in 2016, 2018, 2021, 2022, and 2024 respectively
##2021 SCH? trips should be considered as O trips due to limited info.
##Also I noticed couple travelers begin their day at "O" place. Seems like previous
##trip records are deleted as linked_trip_id is not continuous.
```

```{r clean_sch_probs}
# Here we only focus on 2016/2018 trips as 2021 does not have address info
add_key <- flag_loop_tours %>%
  ungroup() %>%
  unite(
    key, hh_id, day_num, o_lat, departure_time, d_lat, arrival_time,
    remove = FALSE
  )

# check to see how many missing joint trips there are. i.e other hh members
# report hh member 3 was on the trip, but hh member 3 doesn't have that trip
# listed.
temp <- add_key %>%
  select(key, num_hh_travelers) %>%
  group_by(key) %>%
  mutate(count = n()) %>%
  slice(1)
num_missing_joint_trips <- sum(temp$num_hh_travelers) - sum(temp$count)  ##6182 trips
remove(temp)

# this checks to see if anyone in the trip had school as origin/dest place.
# Drop off trips with school trips often take longer (e.g. waiting for the
# bus or carpool).
s_check <- add_key %>%
  select(key, o_place, d_place) %>%
  group_by(key) %>%
  mutate(
    o_s_check = ifelse("S" %in% o_place, 1, 0),
    d_s_check = ifelse("S" %in% d_place, 1, 0)
  ) %>%
  summarize(
    o_s_check = max(o_s_check),
    d_s_check = max(d_s_check)
  )

# Use logic to identify pu/do trips. Many of the SCH? activities are actually
# pu/do.
fix_pudo <- add_key %>%
  left_join(s_check, by = "key") %>%
  group_by(person_id, day_num, tour_num) %>%
  arrange(trip_num, .by_group = TRUE) %>%
  mutate(
    # flag trips where the logic identifies DO before changing purpose
    d_do_flag = case_when(
      d_party_size_change >= 0 ~ 0,
      d_act_dur <= 10 ~ 1,
      d_act_dur <= 25 & d_s_check == 1 ~ 1,
      TRUE ~ 0
    ),
    d_purpose = ifelse(d_do_flag == 1, "PUDO", d_purpose),
    o_purpose = case_when(
      is.na(lag(d_do_flag)) ~ o_purpose,
      lag(d_do_flag) == 1 ~ "PUDO",
      TRUE ~ o_purpose
    ),
    d_pu_flag = case_when(
      d_party_size_change <= 0 ~ 0,
      d_act_dur <= 10 ~ 1,
      d_act_dur <= 25 & o_s_check == 1 ~ 1,
      TRUE ~ 0
    ),
    d_purpose = ifelse(d_pu_flag == 1, "PUDO", d_purpose),
    o_purpose = case_when(
      is.na(lag(d_pu_flag)) ~ o_purpose,
      lag(d_pu_flag) == 1 ~ "PUDO",
      TRUE ~ o_purpose
    ),
    # Differentiate PU/DO at school from other
    d_purpose = ifelse(
      d_purpose == "PUDO" & d_s_check == 1,
      "PUDOS", d_purpose
    ),
    o_purpose = case_when(
      is.na(lag(d_purpose)) ~ o_purpose,
      lag(d_purpose) == "PUDOS" ~ "PUDOS",
      TRUE ~ o_purpose
    )
  ) %>%
  select(-c(o_s_check:d_pu_flag))  

## Update sch? problems
sch_problems <- fix_pudo %>%
  mutate(
    sch_flag = case_when(
      o_purpose == "SCH?" ~ 1,
      d_purpose == "SCH?" ~ 1,
      TRUE ~ 0
    )
  ) %>%
  filter(sch_flag == 1)

num_sch_problems <- sch_problems %>%
  group_by(year) %>%
  summarise(record_count = n())
####401, 124, 32, 75, 25 records in 2016, 2018, 2021, 2022, and 2024 respectively are still SCH?

```

```{r clean_bus_probs}
# School kids going to bus stops are not handled consistently. Fix them.
fix_bus <- fix_pudo %>%
  mutate(
    bus_flag = ifelse(
        mode_type != lead(mode_type) &
        lead(departure_time)-arrival_time <= 15&
        d_purpose %in% c("PreK", "K12", "U") &
        d_place != "S",
      1, 0
    ),
    d_purpose = ifelse(bus_flag == 1, "X", d_purpose),
    o_purpose = case_when(
      is.na(lag(d_purpose)) ~ o_purpose,
      lag(d_purpose) == "X" ~ "X",
      TRUE ~ o_purpose
    )
  ) %>%
  select(-bus_flag, -key)

fix_remaining_sch <- fix_bus %>%
  mutate(
    o_purpose = ifelse(o_purpose == "SCH?" & o_place == "H", "H", o_purpose),
    o_purpose = ifelse(o_purpose == "SCH?", "O", o_purpose),
    d_purpose = ifelse(d_purpose == "SCH?" & d_place == "H", "H", d_purpose),
    d_purpose = ifelse(d_purpose == "SCH?", "O", d_purpose),
  )  
##num_sch_problems=0
```

Numerous problems detected for 2016 and 2018:

- School activities reported by non-students (`r num_sch_problems` trips)
  - Pickup/Dropoff coded as school 
  - Bus stops coded as school rather than mode transfer 
- Missing joint trips (`r num_missing_joint_trips`). E.g., a trip references 
  two hh members but second member doesn't have that trip
- Kids marked as homeschool but go to traditional school on travel day for
  school activity
- Students without school address in person file 
- Home activities reported at non-home locations (e.g. "In-laws house") 
  - This is after we extended the home definition to be 100m around the home (for things like 'front yard' and 'tractor shed')



### Trip linking

**NEED PROSE**
```{r link_out_bus PleaseReview}
## transit access/egress trips are now coded as separate trip records. Need to link them into one trip. Also sometimes people code their walk to bus stop trip not as access/egress
transit_flag <- fix_remaining_sch %>%
  mutate(
    trans_flag = case_when(
      is_access==1 ~ 1, 
      is_egress==1 ~ 1,
      o_purpose == "X" | d_purpose == "X" ~ 1,
      TRUE ~ 0
    )
  )

temp  <- transit_flag  %>%
  filter(trans_flag == 1) %>%
  mutate(
    start_flag = ifelse(o_purpose != "X" & d_purpose == "X", 1, 0),
    end_flag = ifelse(o_purpose == "X" & d_purpose != "X", 1, 0),
    sub_tour = cumsum(start_flag)
  )

# Create a separate df of only the flagged trips. Collapse trip records on
# the same subtour into single trip records.
flagged_only  <- temp  %>%
  ungroup() %>%
  group_by(person_id, day_num, tour_num, sub_tour) %>%
  arrange(trip_num, .by_group = TRUE) %>%
  mutate(
    linked_modes = paste(unique(mode_type), collapse = " "),
    mode_type = "bus",
    across(
      num_hh_travelers:hh_member_9,
      ~ .x[if (all(is.na(duration_minutes))) 1 else which.max(duration_minutes)]
    ),
    trip_weight = mean(trip_weight),
    speed_mph = round(distance_miles / duration_minutes / 60, 0),
    across(
      starts_with(c("destination_", "d_", "arrival_time")),
      ~ last(.x)
    ),
    across(
      c("distance_miles", "duration_minutes", "reported_duration"),
      ~ sum(.x)
    ),
    duration_string = paste0(
      str_pad(duration_minutes %/% 60, 2, "left", "0"),
      ":",
      str_pad(duration_minutes %% 60, 2, "left", "0")
    )
  ) %>%
  slice(1) %>%
  ungroup() %>%
  select(-c(trans_flag:sub_tour)) %>%
  mutate(linked_flag = 1)

xfer_linked_out <- transit_flag %>%
  filter(trans_flag == 0) %>%
  select(-trans_flag) %>%
  bind_rows(flagged_only) %>%
  arrange(year, hh_id, person_id, trip_num) %>%
  relocate(linked_modes, .after = linked_flag) %>%
  mutate(linked_flag = ifelse(is.na(linked_flag), 0, linked_flag))
```




## Exploratory Analysis

TODO: this EDA is using place holder time of day values. Make sure to update
once real values have been determined.

```{r PleaseReview}
prep_for_eda <- xfer_linked_out %>%
  mutate(homebased = ifelse(o_place == "H" | d_place == "H", "HB", "NHB")) %>%
  left_join(
    remove_na_hh %>%
      mutate(child_present = ifelse(num_kids != "0 children", 1, 0)) %>%
      select(
        hh_id, num_people, num_vehicles, num_kids, child_present,
        income_broad
      ),
    by = "hh_id"
  )  %>%
  left_join(
    remove_na_per %>%
      mutate(senior = ifelse(age == "65-74" | age == "75-84" | age == "85 or older", 1, 0)) %>%
      group_by(hh_id) %>%
      summarize(num_seniors = sum(senior)) %>%
      mutate(senior_present = ifelse(num_seniors > 0, 1, 0)),
    by = "hh_id"
  ) %>%
  relocate(num_seniors:senior_present, .after = child_present) %>%
  mutate(
    duration = ifelse(d_act_dur >= 30, "Long", "Short"),
    duration = ifelse(is.na(duration), "Long", duration),
    tod = case_when(
      departure_time < 6*60 ~ "NT",
      departure_time < 9*60 ~ "AM",
      departure_time < (3 + 12)*60 + 30 ~ "MD",
      departure_time < (6 + 12)*60 + 30 ~ "PM",
      TRUE ~ "NT"
    )
  )

trips_purp <- prep_for_eda %>%
  mutate(
    trip_purpose = case_when(
      (!is.na(o_purpose)) & (!is.na(d_purpose)) &
      (o_place != "H") & (d_place != "H") ~ "Other to other", #
      ((o_place == "H") & (d_purpose %in% c("W", "WR", "WRS"))) |
        ((d_place == "H") & (o_purpose %in% c("W", "WR", "WRS"))) ~ "Home to Work", #
      (o_place == "H" & d_purpose %in% c("PreK", "K12")) | 
         (d_place == "H" & o_purpose %in% c("PreK", "K12")) ~ "Home to School",
      (o_place == "H" & d_purpose == "U") | 
         (d_place == "H" & o_purpose== "U") ~ "Home to University",
      (o_place == "H" & d_purpose == "SHP") | 
        (d_place == "H" & o_purpose == "SHP") ~ "Home to Shop", #
      (o_place == "H") & (d_place == "H") ~ "", #
      (o_place == "H" & (!is.na(d_purpose))) | 
        (d_place == "H" & (!is.na(o_purpose))) ~ "Home to Other" #
    )
  )

per_out <- remove_na_per %>% 
  mutate(
    agecat = case_when(
      age == "Under 5"| age == "5-15" | age == "16-17" ~ "Under 18",
      age == "18-24" | age == "25-34" ~ "18-34",
      age == "35-44" | age == "45-54" | age =="55-64" ~ "35-64",
      TRUE ~ "65+"
    )) %>%
  mutate(
    race_simple = ifelse(race_broad == "White", "White", "Non-white")
  )

hh_out <- remove_na_hh %>%
  mutate(
    num_workersalt = ifelse(num_workers>=3, "3+", as.character(num_workers))
  )


```

### Mode simplification
Mode category is quite different in 2021. 2021 survey no longer distinguish between
"rent", "work", or "friend's car" in auto mode - instead, all of them are now in 
"other vehicle" category. 


```{r simplify_mode}
temp <- read_excel(tripconvfile, sheet = "mode_type") 
temp %>%
  rename('Survey Mode' = Translation, 'Simple Mode' = CommLabel) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)


simplify_mode <- trips_purp %>%
  mutate(
    mode_simple = case_when(
      grepl("auto", mode_type) ~ "auto",
#      mode_type == "walk" | mode_type == "bike" ~ "walkbike",
      TRUE ~ mode_type
    ),
    distbin = case_when(
      distance_miles<=1 ~ "Under 1 Mile",
      distance_miles<=6 ~ "Between 1 - 6 Miles",
      TRUE ~ "Over 6 Miles"
    )
  )




```



```{r write tables, eval=FALSE}
# This treatment prevents the output files from being changed unintentionally
# while knitting, testing, etc.
if (1 == 1) {
  write_csv(
    hh_out,
    "C:/Work/Travel Trends/HTS/2024 HTS/processed/hh_processed.csv",
    na = ""
  )
  write_csv(
    per_out,
    "C:/Work/Travel Trends/HTS/2024 HTS/processed/per_processed.csv",
    na = ""
  )
  write_csv(
    simplify_mode,
    "C:/Work/Travel Trends/HTS/2024 HTS/processed/trips_processed.csv",
    na = ""
  )
  write_csv(
    remove_na_day,
    "C:/Work/Travel Trends/HTS/2024 HTS/processed/day_processed.csv",
    na = ""
  )
}
```

